---
  title: "ITU risk scoring analysis by ethnicity"
author: "Dr C J Martin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document: default
pdf_document: default
---
  
  This R code analyses data extracted from the eICU and MIMIC-III database of individual patient clinicla data from ITUs in the USA.
It was accessed from www.physionet.org as part of an approved project in association with the following authors:
  
  Rahuldeb Sarkar(1,2), Christopher Martin(3,4), Heather Mattie(5), Judy Wawira Gichoya(6), David J. Stone(7), Leo Anthony Celi(8,9,10)

1. Departments of Respiratory Medicine and Critical Care, Medway NHS Foundation Trust, Gillingham, Kent, UK; 
2. Faculty of Life Sciences, King’s College London, London, UK
3. UCL Institute for Health Informatics, London, UK; 
4. Crystallise Ltd, UK.
5. Department of Biostatistics, Harvard T.H. Chan School of Public Health, Boston, MA, USA, 02115
6. Interventional Radiology & Informatics, Department of Radiology & Imaging Sciences, Emory University, 1364 Clifton Rd NE Suite AG08 Atlanta, GA 30322
7. Departments of Anesthesiology and Neurosurgery, and the Center for Advanced Medical Analytics, University of Virginia School of Medicine, Charlottesville, VA, 22908   
8. Laboratory for Computational Physiology, Massachusetts Institute of Technology, Cambridge, MA, USA 20139; 9Division of Pulmonary, Critical Care and Sleep Medicine, Beth Israel Deaconess Medical Center, Boston, MA, USA 02215; 10Department of Biostatistics, Harvard T.H. Chan School of Public Health, Boston, MA, USA, 02115



```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr) # ntile function.
library(knitr)  # Rmarkdown.
library(rsq)   # To calculate the partial R2 from the logistic regression.
library(ems) # For the SMR function.
library(gbm) # calibrate.plot function
library(pROC) # For generating ROC curves (roc function).
library(reshape2) # Used for 'dcast' function.

source("calibrationPlot.r")

# Parameters for graphic output.
figWidth <- 7
figHeight <- 6
figWidth2 <- 5
figHeight2 <- 4

knitr::opts_chunk$set(out.width = "50%")
options(OutDec="\xB7")
```


```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=12, fig.height=8}
##If the package 'forestplot' is not installed then Rstudio should prompt 
##you to install it. If not then remove the hashtag from the line below
##before running the script and it will be installed.
#############################################################################################

plotForest <- function(theData, Vhispanic, Vblack, Vwhite, Vasian, Vall, theScore, theMetric, doPNG, theCounts, tag ){
  library(forestplot)
  
  options(OutDec = ".")
  
  source<-c("Ethnic group",c( "Hispanic","Black", "White","Asian","All"))
  year<-c("Year",c("0","0","0","0","0"))
  numberInSample <- theCounts
  # measure<-c("Data set",rep(theData, 5))
  value<-c("Value",Vhispanic[2],Vblack[2],Vwhite[2],Vasian[2],Vall[2])
  isSummary<-c(TRUE,FALSE,FALSE,FALSE,FALSE,TRUE)
  lowerCI<-c(NA,as.numeric(c(Vhispanic[1],Vblack[1],Vwhite[1],Vasian[1],Vall[1])))
  upperCI<-c(NA,as.numeric(c(Vhispanic[3],Vblack[3],Vwhite[3],Vasian[3],Vall[3])))
  
  ## Define how the plot is to look.
  
  # Define the colours for the forest plot boxes, lines, and summary prisms.
  boxColour='black' #"#008080"
  lineColour='black' #"#008080"
  summaryColour='black' #"#008080"
  
  # Define the size of the plot boxes.
  boxsize=0.2
  
  # Define the dimensions of the plot.
  plotWidth=900
  plotHeight=500
  
  # Define the upper and lower bounds of the x-axis.
  lowerBound=floor(min(lowerCI[2:6])*100)/100
  upperBound=ceiling(max(upperCI[2:6])*100)/100
  
  # Define the x-axis tick marks: the lower bound, upper bound, and interval
  lowerTicks=floor(min(lowerCI[2:6])*100)/100
  upperTicks=ceiling(max(upperCI[2:6])*100)/100
  ticksInterval=0.02
  ticksSize=1.3
  
  # Define the x-axis label and text size.
  xlab= theMetric
  xlabSize=1.3
  
  # Define what paper details to include in addition to the source.
  include_year=FALSE
  include_numberInSample=TRUE
  include_value=TRUE
  
  ##Create data for plot.
  tableText<-c(source)
  
  if (include_year==TRUE){
    tableText<-cbind(tableText,year)
  }
  if (include_numberInSample==TRUE){
    tableText<-cbind(tableText,numberInSample)
  }
  
  if (include_value==TRUE){
    tableText<-cbind(tableText,c(paste(theMetric,"(95% CI)"),paste(value[2:6], " (",lowerCI[2:6],", ",upperCI[2:6],")", sep="")))
  }
  
  tableData<-structure(list(value,lowerCI,upperCI),class="data.frame")
  
  ##Plot the forest plot
  xClip=c(lowerBound,upperBound)
  own<-fpTxtGp(gpar(cex=1.3))
  own$ticks$cex<-ticksSize
  own$xlab$cex<-xlabSize
  
  if(doPNG){png(paste0("ForestPlot_output_",theData,"_",theMetric,"_",tag,".png"),width=plotWidth, height=plotHeight,unit="px")}
  
  this <- getOption("OutDec")
  forestplot(tableText,
             mean=value,
             lower=lowerCI,
             upper=upperCI,
             title=paste(theMetric, "for", theScore, "in each ethnic group \nin the", theData, "data"),
             is.summary=isSummary,
             grid=1,
             col=fpColors(box=boxColour,line=lineColour,summary=summaryColour),
             zero=NA,
             txt_gp=own,
             clip=xClip,
             xlab=xlab,
             xticks = seq(from=lowerTicks, to = upperTicks, by = ticksInterval),
             hrzl_lines=TRUE,
             vertices=TRUE,
             boxsize=boxsize,
             graphwidth = "auto"
  )
  options(OutDec="\xB7")
  
  if(doPNG){dev.off()}
}
```



```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, fig.height = figHeight2, fig.width = figWidth2}

logisticPlot <- function(binWidth = 1, database="MIMIC-III", scoreType="SOFA", ethList=c("White","Black","Hispanic","Asian"), colours=c("All"='black',"Hispanic"='red', "Black"='blue', "White"='seagreen4', "Asian"='orange'), ethNames = c(hisp = "Hispanic", black = "Black", white = "White", asian = "Asian"), mode = 'logistic'){
  
  if(database == "eICU"){
    db <- eICU_Dat
    ethnicGroup <- eICU_Dat$ethnicity
    deathField <- "DiedHosp"
    if(scoreType == "SOFA"){
      theScore <- eICU_Dat$SOFA1
    }else{
      theScore <- eICU_Dat$apachescore
    }
  }else{
    db <- icustays
    ethnicGroup <- icustays$ETHNICITY2
    deathField <- "DiedHosp"
    if(scoreType == "SOFA"){
      theScore <- icustays$SOFA_Adm
    }else{
      theScore <- icustays$OASIS
    }
  }
  
  db$scoreBin <- theScore %/% binWidth
  
  # Set up a table of weights for the fitting.
  scoreMax <- max(db$scoreBin)
  weightsList <- data.frame(matrix(rep(0, scoreMax+1*6), nrow = scoreMax+1, ncol = 6))
  weightsList[,1] <- 0:scoreMax
  colnames(weightsList) <- c("score","ALL",ethList)
  theseWeights <- table(db$scoreBin)
  theseWeights <- data.frame(cbind(as.integer(rownames(theseWeights)), as.integer(theseWeights)))
  labels <- match(weightsList$score, theseWeights[,1])
  weightsList[,2] <- theseWeights[match(weightsList$score, theseWeights[,1]),2]
  weightsList[is.na(weightsList[,2]),] <- 0
  
  for(eth in ethList){
    theseWeights <- table(db[ethnicGroup == eth,]$scoreBin)
    theseWeights <- data.frame(cbind(as.integer(rownames(theseWeights)), as.integer(theseWeights)))
    weightsList[,eth] <- theseWeights[match(weightsList$score, theseWeights[,1]),2]
    weightsList[is.na(weightsList[,eth]),eth] <- 0
  }
  
  deathByScore_T <- aggregate(db[, deathField], list(db$scoreBin), mean)
  deathByScore <- data.frame(matrix(nrow = scoreMax+1, ncol = 2))
  colnames(deathByScore) <- c("score","All")
  deathByScore$score <- 0:scoreMax
  deathByScore[,2] <- deathByScore_T[match(deathByScore$score, deathByScore_T[,1]),2]
  for(eth in ethList){
    deathByScore_T <- aggregate(db[ethnicGroup == eth, deathField], list(db[ethnicGroup == eth, ]$scoreBin), mean)
    deathByScore[,eth] <- cbind(deathByScore, rep(NA, scoreMax+1))
    deathByScore[,eth] <- deathByScore_T[match(deathByScore$score, deathByScore_T[,1]),2]
  }
  
  if(mode == 'logistic'){
    fitted <- predict(glm(deathByScore$All ~ deathByScore$score, family = 'binomial', weights = weightsList$ALL), data.frame(score = 0:scoreMax), 
                      type = "response", se.fit = TRUE)
    lineData <- fitted$fit
    
    SOFAmort <- NA
    if(database == "eICU"){
      if(scoreType == "SOFA"){
        SOFAmort <- fitted$fit[eICU_Dat$SOFA1+1]
        write.csv(SOFAmort, "SOFAmort_eICU.csv")
      }else{
        
      }
    }else{
      if(scoreType == "SOFA"){
        SOFAmort <- fitted$fit[icustays$SOFA_Adm+1]
        write.csv(SOFAmort, "SOFAmort_MIMIC.csv")
      }else{
        
      }
    }
  }else{
    deathByScore$All[is.na(deathByScore$All)] <- 0
    lineData <- biviso(y=as.matrix(cbind(deathByScore$All,deathByScore$score)))[,1]
  }
  plot(y=lineData, x=0:scoreMax, type = 'l', col=colours["All"], lwd=2, main=paste0("Plot of ", mode," curve fitted to \nobserved mortality by ", scoreType, " score in ", database), xlab = paste(scoreType, "score"), ylab = "Mortality rate" , cex.main = 1 )
  if(mode == 'logistic'){
    polygon(c(0:scoreMax, scoreMax:0), c(fitted$fit-1.96*fitted$se.fit, rev(fitted$fit+1.96*fitted$se.fit)),   col = adjustcolor(colours["All"], alpha=0.05), border = NA)
  }
  if(mode == 'logistic'){
    fitted <- predict(glm(as.matrix(deathByScore[ethNames["white"]]) ~ deathByScore$score, family = 'binomial', weights = as.matrix(weightsList[,ethNames["white"]])), data.frame(score = 0:scoreMax), type = "response", se.fit = TRUE)
    lineData <- fitted$fit
  }else{
    deathByScore[,ethNames["white"]][is.na(deathByScore[,ethNames["white"]])] <- 0
    lineData <- biviso(y=as.matrix(cbind(deathByScore[,ethNames["white"]],deathByScore$score)))[,1]
  }
  lines(y=deathByScore[,ethNames["white"]], x = 0:scoreMax, col=colours["White"], type = 'p', pch = 20, cex = 0.75)
  lines(y=lineData, x=0:scoreMax, type = 'l', col=colours["White"] , lwd=2)
  if(mode == 'logistic'){
    polygon(c(0:scoreMax, scoreMax:0), c(fitted$fit-1.96*fitted$se.fit, rev(fitted$fit+1.96*fitted$se.fit)),   col = adjustcolor(colours["White"], alpha=0.2), border = NA)
  }
  if(mode == 'logistic'){
    fitted <- predict(glm(as.matrix(deathByScore[ethNames["asian"]]) ~ deathByScore$score, family = 'binomial', weights = as.matrix(weightsList[ethNames["asian"]])), data.frame(score = 0:scoreMax), type = "response", se.fit = TRUE)  #
    lineData <- fitted$fit
  }else{
    deathByScore[,ethNames["asian"]][is.na(deathByScore[,ethNames["asian"]])] <- 0
    lineData <- biviso(y=as.matrix(cbind(deathByScore[,ethNames["asian"]],deathByScore$score)))[,1]
  }
  lines(y=deathByScore[,ethNames["asian"]], x = 0:scoreMax, col=colours["Asian"], type = 'p', pch = 20, cex = 0.75)
  lines(y=lineData, x=0:scoreMax, type = 'l', col=colours["Asian"] , lwd=2)
  if(mode == 'logistic'){
    polygon(c(0:scoreMax, scoreMax:0), c(fitted$fit-1.96*fitted$se.fit, rev(fitted$fit+1.96*fitted$se.fit)),   col = adjustcolor(colours["Asian"], alpha=0.2), border = NA)
  }
  if(mode == 'logistic'){
    fitted <- predict(glm(as.matrix(deathByScore[ethNames["black"]]) ~ deathByScore$score, family = 'binomial', weights = as.matrix(weightsList[ethNames["black"]])), data.frame(score = 0:scoreMax), type = "response", se.fit = TRUE)
    lineData <- fitted$fit
  }else{
    deathByScore[ethNames["black"]][is.na(deathByScore[ethNames["black"]])] <- 0
    lineData <- biviso(y=as.matrix(cbind(deathByScore[ethNames["black"]],deathByScore$score)))[,1]
  }
  lines(y=deathByScore[,ethNames["black"]], x = 0:scoreMax, col=colours["Black"], type = 'p', pch = 20, cex = 0.75)
  lines(y=lineData, x=0:scoreMax, type = 'l', col=colours["Black"], lwd=2)
  if(mode == 'logistic'){
    polygon(c(0:scoreMax, scoreMax:0), c(fitted$fit-1.96*fitted$se.fit, rev(fitted$fit+1.96*fitted$se.fit)),   col = adjustcolor(colours["Black"], alpha=0.2), border = NA)
  }
  if(mode == 'logistic'){
    fitted <- predict(glm(as.matrix(deathByScore[ethNames["hisp"]]) ~ deathByScore$score, family = 'binomial', weights = as.matrix(weightsList[ethNames["hisp"]])), data.frame(score = 0:scoreMax), type = "response", se.fit = TRUE)
    lineData <- fitted$fit
  }else{
    deathByScore[ethNames["hisp"]][is.na(deathByScore[ethNames["hisp"]])] <- 0
    lineData <- biviso(y=as.matrix(cbind(deathByScore[ethNames["hisp"]],deathByScore$score)))[,1]
  }
  lines(y=deathByScore[,ethNames["hisp"]], x = 0:scoreMax, col=colours["Hispanic"], type = 'p', pch = 20, cex = 0.75)
  lines(y=lineData, x=0:scoreMax, type = 'l', col=colours["Hispanic"], lwd =2 )
  if(mode == 'logistic'){
    polygon(c(0:scoreMax, scoreMax:0), c(fitted$fit-1.96*fitted$se.fit, rev(fitted$fit+1.96*fitted$se.fit)),   col = adjustcolor(colours["Hispanic"], alpha=0.2), border = NA)
  }
  legend("topleft", legend = c("All", "Hispanic", "Black", "White", "Asian" ), col = c('black', 'red', 'blue', 'seagreen4', 'orange'), pch = c(19, 19, 19, 19, 19), cex = 0.8 )
  #dev.off()
  
}
```



```{r eval = TRUE, echo = FALSE, message = FALSE, warning =FALSE}

# Set the age range of interest.
minAge <- 16
maxAge <- 90

# Set the list of ethnicity names. They are recorded differnently in the two databases
ethnicitiesList <- c("White","Black","Hispanic","Asian")

# These are the data files from physionet.org via the Google BigQuery platform.
# The data files cannot be supplied with this code as it requires approved access to the MIMIC-III and eICU database.
# SQL code was used to create and extract SOFA scores.
# This is the 
SOFA <- read.csv(file="bq-results-20200805-214538-n690szu0x80_SOFAscoresAdmission_DUMMY.csv", stringsAsFactors = FALSE)
OASIS <- read.csv(file="bq-results-20200806-202359-anbq1d2oo30vOASIS_DUMMY.csv", stringsAsFactors = FALSE)
patients <- read.csv(file="bq-results-20200806-225103-mcsk0rb8h76l_Patients_DUMMY.csv", stringsAsFactors = FALSE)
admissions <- read.csv(file="bq-results-20200806-231458-u3iow35qz30w_Admissions_DUMMY.csv", stringsAsFactors = FALSE)
icustays <- read.csv(file = "bq-results-20200808-092553-ttzoex49foff_ICUSTAYS_DUMMY.csv")
DoBgender <- read.csv("bq-results-20200806-082126-2pfodnem4miv_DOBgender_DUMMY.csv")
ethnicitiesM <- read.csv("bq-results-20200806-081503-l2nmx172wnn4_SubjectsEthnicities_DUMMY.csv")

admissions$DiedHosp <- !(admissions$DEATHTIME == "")

patients$ETHNICITY <- ethnicitiesM$ETHNICITY[match(patients$SUBJECT_ID, ethnicitiesM$SUBJECT_ID)]
patients <- patients[,c("SUBJECT_ID","GENDER","DOB","ETHNICITY")]

MIMICnos <- list()
MIMICnos$Total <- nrow(icustays)

icustays$INTIME <- as.Date(icustays$INTIME)
icustays$DOB <- patients$DOB[match(icustays$SUBJECT_ID, patients$SUBJECT_ID)]
icustays$DOB <- as.Date(icustays$DOB)
icustays$AGE <- (as.Date(icustays$INTIME) - as.Date(icustays$DOB))/365.25
#icustays$AGE <- as.Date(icustays$INTIME) - as.Date(icustays$DOB)
icustays$OASIS <- OASIS$oasis[match(icustays$ICUSTAY_ID,OASIS$icustay_id)]
icustays$OASISprob <- OASIS$oasis_PROB[match(icustays$ICUSTAY_ID,OASIS$icustay_id)]
icustays[,c("GENDER","DOB","ETHNICITY")] <- patients[match(icustays$SUBJECT_ID,patients$SUBJECT_ID),c("GENDER","DOB","ETHNICITY")]
icustays$SOFA_Adm <- SOFA$SOFA[match(icustays$ICUSTAY_ID,SOFA$icustay_id)]
icustays$DiedHosp <- admissions$DiedHosp[match(icustays$HADM_ID, admissions$HADM_ID)]
icustays$INTIME <- as.Date(icustays$INTIME)

MIMICnos$originalCheck <- nrow(icustays)

# Drop any under the minimum age or over the maximum age.
MIMICnos$agedMin <- nrow(icustays[icustays$AGE <= minAge,])
MIMICnos$agedMax <- nrow(icustays[icustays$AGE >= maxAge,])
MIMICnos$ageOut <- MIMICnos$agedMin + MIMICnos$agedMax
icustays <- icustays[icustays$AGE >= minAge & icustays$AGE <= maxAge,]

MIMICnos$exAges <- nrow(icustays)
MIMICnos$episodeNos <- nrow(ethnicitiesM)
MIMICnos$ethMissUnable <- nrow(ethnicitiesM[ethnicitiesM$ETHNICITY == "UNABLE TO OBTAIN",])
MIMICnos$ethMissUnknown <- nrow(ethnicitiesM[ethnicitiesM$ETHNICITY == "PATIENT DECLINED TO ANSWER",])
MIMICnos$ethMissDeclined <- nrow(ethnicitiesM[ethnicitiesM$ETHNICITY == "UNKNOWN/NOT SPECIFIED",])
MIMICnos$NatAm <- nrow(icustays[grepl("INDIAN", icustays$ETHNICITY),])
MIMICnos$ethMissing <- MIMICnos$ethMissDeclined + MIMICnos$ethMissUnable + MIMICnos$ethMissUnknown + MIMICnos$NatAm
MIMICnos$missPredMort <- nrow(icustays[is.na(icustays$OASISprob) | icustays$OASISprob == "",])

MIMICnos$Miss_percent <- MIMICnos$ethMissing/nrow(ethnicitiesM)*100

```

# MIMIC-III
## Data cleaning
There was a total of `r MIMICnos$Total` ICU episodes in the database.  
There were `r MIMICnos$agedMin` episodes aged under `r minAge` who were excluded (`r round(MIMICnos$agedMin/MIMICnos$Total*100,1)`%). 
There were `r MIMICnos$agedMax` episodes  (`r round(MIMICnos$agedMax/MIMICnos$Total*100,1)`%) with an incorrect date of birth who were excluded as the OASIS score uses age as a rating factor.

Of the `r MIMICnos$episodeNos` admissions included after removing those under `r minAge` or over `r maxAge`:  
  `r MIMICnos$ethMissing` did not have a recording of their ethnicity (`r round(MIMICnos$Miss_percent,1)`%);  
`r MIMICnos$missPredMort` had no valid predicted mortality (`r round(MIMICnos$missPredMort/MIMICnos$episodeNos*100,1)`%).

Next we reclassify the various ethnic groups according to the broader categories that we are interested in: "White", "Black", "Asian", and "Hispanic".
We then drop ambiguous ethnic groups such as "Portugese", "South American" and "MULTI RACE ETHNICITY", and the "Native American" as there are inadequate numbers. 

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

ethnicitiesM$ETHNICITY2 <- NA
for(eth in ethnicitiesList){
  ethnicitiesM[grepl(eth,ethnicitiesM$ETHNICITY),"ETHNICITY2"] <- eth
}
MIMICnos$notKeyEthnicity <- nrow(ethnicitiesM[is.na(ethnicitiesM$ETHNICITY2),])

icustays$ETHNICITY2 <- NA
for(eth in ethnicitiesList){
  icustays[grepl(toupper(eth),icustays$ETHNICITY),"ETHNICITY2"] <- eth
}
MIMICnos$ethOutScope <- nrow(icustays[is.na(icustays$ETHNICITY2),])
icustays <- icustays[icustays$ETHNICITY2 %in% ethnicitiesList,]

MIMICnos$incEthnicity1 <- nrow(icustays)

outTable <- rbind(data.frame(table(icustays$ETHNICITY2)), sum(table(icustays$ETHNICITY2)))
outTable[,1] <- c(rev(ethnicitiesList), "ALL")

knitr::kable(outTable, align=c('r','c'), caption="Numbers of episodes by ethnic group (MIMIC-III).")

```

`r MIMICnos$NatAm` (`r round(MIMICnos$NatAm/MIMICnos$Total*100,1)`%) were Native American.

Altogether, there were `r MIMICnos$ethOutScope` admissions that were not identified as being in the target group of 4 ethnicities (`r round(MIMICnos$ethOutScope/MIMICnos$Total,3)*100`%).


Of these, `r MIMICnos$ethMissing` (`r round(MIMICnos$ethMissing/MIMICnos$Total,3)*100`%) had no recorded ethnicity at all.

\pagebreak
# Baseline characteristics

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

baseTable1 <- data.frame(matrix(ncol=7, nrow=(length(ethnicitiesList)+1)))
rownames(baseTable1) <- c(ethnicitiesList, "All included")
for(eth in ethnicitiesList){
  icustaysX <- icustays[icustays$ETHNICITY2 == eth,]
  icustaysX$GENDER[icustaysX$GENDER=="M"] <- 0
  icustaysX$GENDER[icustaysX$GENDER=="F"] <- 1
  icustaysX$DiedHosp[icustaysX$DiedHosp == "FALSE"] <- 0
  icustaysX$DiedHosp[icustaysX$DiedHosp == "TRUE"] <- 1
  NewRow <- as.matrix(t(c(nrow(icustaysX), round(mean(as.numeric(icustaysX$AGE), na.rm=TRUE),1), paste(round(quantile(icustaysX$AGE, 0.25, na.rm=TRUE)[[1]],1), "-",round(quantile(icustaysX$AGE, 0.75, na.rm=TRUE)[[1]]),1), round(mean(as.numeric(icustaysX$GENDER), na.rm=TRUE),2), median(icustaysX$OASIS, na.rm=TRUE), paste(quantile(icustaysX$OASIS, 0.25)[[1]], "-",quantile(icustaysX$OASIS, 0.75)[[1]]), round(mean(icustaysX$OASISprob, na.rm=TRUE),4))))
  baseTable1[rownames(baseTable1) == eth,] <- NewRow
}
colnames(baseTable1) <- c("Num", "median age","IQR age", "proportion female","median OASIS score", "IQR OASIS score", "mean predicted hospital mortality")

# Now add all the included ethnicities together.
icustaysX <- icustays[icustays$ETHNICITY2 %in% ethnicitiesList,]
MIMICnos$incEthnicity2 <- nrow(icustaysX)
icustaysX$GENDER[icustaysX$GENDER=="M"] <- 0
icustaysX$GENDER[icustaysX$GENDER=="F"] <- 1
icustaysX$DiedHosp[icustaysX$DiedHosp == "FALSE"] <- 0
icustaysX$DiedHosp[icustaysX$DiedHosp == "TRUE"] <- 1
NewRow <- as.matrix(t(c(nrow(icustaysX), round(median(as.numeric(icustaysX$AGE), na.rm=TRUE),1), paste(round(quantile(icustaysX$AGE, 0.25, na.rm=TRUE)[[1]],1), "-",round(quantile(icustaysX$AGE, 0.75, na.rm=TRUE)[[1]],1)), round(mean(as.numeric(icustaysX$GENDER), na.rm=TRUE),2), median(icustaysX$OASIS, na.rm=TRUE), paste(quantile(icustaysX$OASIS, 0.25)[[1]], "-",quantile(icustaysX$OASIS, 0.75)[[1]]), round(mean(icustaysX$OASISprob, na.rm=TRUE),4))))
baseTable1[rownames(baseTable1) == "All included",] <- NewRow



baseTable2 <- data.frame(matrix(ncol=4, nrow=(length(ethnicitiesList)+1)))
ethnicities <- c("White","Black","Hispanic","Asian","NATIVE AMERICAN")
rownames(baseTable2) <- c(ethnicitiesList, "All included")
for(eth in ethnicities){
  icustaysX <- icustays[icustays$ETHNICITY2 == eth & icustays$AGE >= minAge,]
  icustaysX$GENDER[icustaysX$GENDER=="M"] <- 0
  icustaysX$GENDER[icustaysX$GENDER=="F"] <- 1
  icustaysX$DiedHosp[icustaysX$DiedHosp == "FALSE"] <- 0
  icustaysX$DiedHosp[icustaysX$DiedHosp == "TRUE"] <- 1
  NewRow <- as.matrix(t(c(round(sd(icustaysX$OASISprob, na.rm=TRUE),4), round(mean(icustaysX$DiedHosp, na.rm=TRUE),3), round(median(icustaysX$LOS, na.rm = TRUE),1), paste(round(quantile(icustaysX$LOS, 0.25, na.rm=TRUE)[[1]],1), "-",round(quantile(icustaysX$LOS, 0.75, na.rm=TRUE[[1]]),1)))))
  baseTable2[rownames(baseTable2) == eth,] <- NewRow
}
colnames(baseTable2) <- c("sd predicted hospital mortality", "hospital mortality","median length of stay", "IQR for LoS")

# Now add all the included ethnicities together.
icustaysX <- icustays[icustays$ETHNICITY2 %in% ethnicitiesList,]
MIMICnos$incEthnicity2 <- nrow(icustaysX)
icustaysX$GENDER[icustaysX$GENDER=="M"] <- 0
icustaysX$GENDER[icustaysX$GENDER=="F"] <- 1
icustaysX$DiedHosp[icustaysX$DiedHosp == "FALSE"] <- 0
icustaysX$DiedHosp[icustaysX$DiedHosp == "TRUE"] <- 1
NewRow <- as.matrix(t(c(round(sd(icustays$OASISprob, na.rm=TRUE),4), round(mean(icustays$DiedHosp, na.rm=TRUE),3), round(median(icustays$LOS, na.rm = TRUE),1), paste(round(quantile(icustays$LOS, 0.25, na.rm=TRUE)[[1]],1), "-",round(quantile(icustays$LOS, 0.75, na.rm=TRUE[[1]]),1)))))
baseTable2[rownames(baseTable1) == "All included",] <- NewRow

knitr::kable(baseTable1, format.args = c(width=70), caption = "Baseline patient characteristics in MIMIC-III.")

```


\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(baseTable2, format.args = c(width=70), caption = "Baseline patient characteristics in MIMIC-III cont'd.")

```

\pagebreak

# Output linear models of ethnicity as a predictor of each of the baseline characteristics to see if the vary signifiacntl by ethnicity.

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}


pAgeX <- lm(as.numeric(icustays$AGE) ~ icustays$ETHNICITY2)
summary(pAgeX)
icustays$GENDER2 <- 0
icustays$GENDER2[icustays$GENDER == "F"] <- 1
pGender <- lm(icustays$GENDER2 ~ icustays$ETHNICITY2)
summary(pGender)
```

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
pOasis <- lm(icustays$OASIS ~ icustays$ETHNICITY2)
summary(pOasis)

pPredMort <- lm(icustays$OASISprob ~ icustays$ETHNICITY2)
summary(pPredMort)
```

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
pHospMort <- lm(icustays$DiedHosp ~ icustays$ETHNICITY2)
summary(pHospMort)
pLoS <- lm(icustays$LOS ~ icustays$ETHNICITY2)
summary(pLoS)
```

\pagebreak
# Calibration plots
The total number of episodes included in the analysis is `r nrow(icustays)`.

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width=figWidth2, fig.height=figHeight2}
diedProp <- aggregate(icustays$DiedHosp, list(icustays$ETHNICITY2),mean)

icustays$OAISISdecile <- ntile(as.numeric(icustays$OASISprob), 10)

deathByOasisDecile <- aggregate(icustays$DiedHosp, list(icustays$OAISISdecile), mean)
OasisPbyOasisDecile <- aggregate(icustays$OASISprob, list(icustays$OAISISdecile), mean)
plot(deathByOasisDecile[,2], OasisPbyOasisDecile[,2], main="This is a plot of the proportion of deaths in each \ndecile of Oasis predicted mortality (MIMIC-III).", xlab="Oasis predicted mortality", ylab = "Actual mortality", cex.main=0.9)
OASISprobCor <- cor(deathByOasisDecile[,2], OasisPbyOasisDecile[,2])
```
```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

#par(bg='white', lwd=2, col = 'black')
calibrate.plot_2(icustays$DiedHosp, icustays$OASISprob, line.par = list(col='black'), shade.col = adjustcolor('gray', alpha = 0.1), main="Calibration plots for OASIS in MIMIC-III")
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "White",]$DiedHosp, icustays[icustays$ETHNICITY2 == "White",]$OASISprob, replace = FALSE, line.par = list(lwd=2, col='seagreen4'), shade.col = adjustcolor('seagreen4', alpha = 0.1))
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Asian",]$OASISprob, replace = FALSE, line.par = list(lwd=2, col='orange'), shade.col = adjustcolor('orange', alpha = 0.1))
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Black",]$OASISprob, replace = FALSE, line.par = list(lwd=2, col='blue'), shade.col = adjustcolor('blue', alpha = 0.1))
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Hispanic",]$OASISprob, replace = FALSE, line.par = list(lwd=2, col='red'), shade.col = adjustcolor('red', alpha = 0.1))
legend('topleft',legend=c("All","Hispanic","Black","White","Asian"), lwd=c(2,2,2,2,2), col=c('black','red','blue','seagreen4','orange'))

```



```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
# MIMIC Oasis
SMRall_MIMIC_Oa <- SMR(as.numeric(icustays$DiedHosp), icustays$OASISprob)
SMRall_MIMIC_Oa <- c(SMRall_MIMIC_Oa[5], SMRall_MIMIC_Oa[4], SMRall_MIMIC_Oa[6])
SMRasian_MIMIC_Oa <- SMR(as.numeric(icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp), icustays[icustays$ETHNICITY2 == "Asian",]$OASISprob)
SMRasian_MIMIC_Oa <- c(SMRasian_MIMIC_Oa[5], SMRasian_MIMIC_Oa[4], SMRasian_MIMIC_Oa[6])
SMRblack_MIMIC_Oa <- SMR(as.numeric(icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp), icustays[icustays$ETHNICITY2 == "Black",]$OASISprob)
SMRblack_MIMIC_Oa <- c(SMRblack_MIMIC_Oa[5], SMRblack_MIMIC_Oa[4], SMRblack_MIMIC_Oa[6])
SMRhispanic_MIMIC_Oa <- SMR(as.numeric(icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp), icustays[icustays$ETHNICITY2 == "Hispanic",]$OASISprob)
SMRhispanic_MIMIC_Oa <- c(SMRhispanic_MIMIC_Oa[5], SMRhispanic_MIMIC_Oa[4], SMRhispanic_MIMIC_Oa[6])
SMRwhite_MIMIC_Oa <- SMR(as.numeric(icustays[icustays$ETHNICITY2 == "White",]$DiedHosp), icustays[icustays$ETHNICITY2 == "White",]$OASISprob)
SMRwhite_MIMIC_Oa <- c(SMRwhite_MIMIC_Oa[5], SMRwhite_MIMIC_Oa[4], SMRwhite_MIMIC_Oa[6])

```

# Recalibration of mortality estimates to OASIS scores.

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, fig.width=figWidth2, fig.height=figHeight2}
plotColours <- c("All"='black',"Hispanic"='red', "Black"='blue', "White"='seagreen4', "Asian"='orange')
logisticPlot(binWidth = 1, database = "MIMIC-III", scoreType = "Oasis", ethnicitiesList, colours=plotColours, ethNames = c(hisp = "Hispanic", black = "Black", white = "White", asian = "Asian"), 'logistic')    # modes: 'logistic' or 'isotonic regression'.
```


```{r eval = TRUE, echo = FALSE, message = FALSE, include=FALSE, warning = FALSE}
# Calculate the ROCs for Oasis
library(ROCR)
thePredictionsAll_M <-  icustays$OASISprob
thePredictionsAsian_M <- icustays[icustays$ETHNICITY2 == "Asian",]$OASISprob
thePredictionsBlack_M <- icustays[icustays$ETHNICITY2 == "Black",]$OASISprob
thePredictionsHispanic_M <- icustays[icustays$ETHNICITY2 == "Hispanic",]$OASISprob
thePredictionsWhite_M <- icustays[icustays$ETHNICITY2 == "White",]$OASISprob

theLabelsAll_M <- icustays$DiedHosp
theLabelsAsian_M <- icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp
theLabelsBlack_M <- icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp
theLabelsHispanic_M <- icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp
theLabelsWhite_M <- icustays[icustays$ETHNICITY2 == "White",]$DiedHosp

predObjAll_M <- prediction(predictions = thePredictionsAll_M, labels = theLabelsAll_M)
aucAll_M <- round(performance(predObjAll_M, measure = "auc")@y.values[[1]],3)
perfObjAll_M <- performance(predObjAll_M, "tpr","fpr")
predObjAsian_M <- prediction(predictions = thePredictionsAsian_M, labels = theLabelsAsian_M)
aucAsian_M <- round(performance(predObjAsian_M, measure = "auc")@y.values[[1]],3)
perfObjAsian_M <- performance(predObjAsian_M, "tpr","fpr")
predObjBlack_M <- prediction(predictions = thePredictionsBlack_M, labels = theLabelsBlack_M)
aucBlack_M <- round(performance(predObjBlack_M, measure = "auc")@y.values[[1]],3)
perfObjBlack_M <- performance(predObjBlack_M, "tpr","fpr")
predObjHispanic_M <- prediction(predictions = thePredictionsHispanic_M, labels = theLabelsHispanic_M)
aucHispanic_M <- round(performance(predObjHispanic_M, measure = "auc")@y.values[[1]],3)
perfObjHispanic_M <- performance(predObjHispanic_M, "tpr","fpr")
predObjWhite_M <- prediction(predictions = thePredictionsWhite_M, labels = theLabelsWhite_M)
aucWhite_M <- round(performance(predObjWhite_M, measure = "auc")@y.values[[1]],3)
perfObjWhite_M <- performance(predObjWhite_M, "tpr","fpr")

```

# Discrimination of OASIS

```{r eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE, fig.width=figWidth2, fig.height=figHeight2}
{
  plot(perfObjAll_M, main = "ROC curves for the Oasis predicted hospital mortality \nin the MIMIC-III database by ethnicity.", lwd=3, cex.main=0.9)
  plot(perfObjAsian_M, add = TRUE, col = 'orange', lwd=3)
  plot(perfObjBlack_M, add = TRUE, col = 'blue', lwd=3)
  plot(perfObjHispanic_M, add = TRUE, col = 'red', lwd=3)
  plot(perfObjWhite_M, add = TRUE, col = 'seagreen4', lwd=3)
  abline(0,1,col='black', lty=2)
  op <- par(cex = 0.8)
  legend('bottomright',legend = c(paste("All", aucAll_M),paste("Hispanic",aucHispanic_M),paste("Black", aucBlack_M),  paste("White",aucWhite_M),paste("Asian", aucAsian_M)), col=c("black","red","blue","seagreen4","orange"), lty = c(1,1,1,1,1), lwd=c(3,3,3,3,3))
  
}

```

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

AUCall_Oasis <- round(ci.auc(icustays$DiedHosp, icustays$OASISprob),3)
AUCasian_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Asian",]$OASISprob),3)
AUCblack_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Black",]$OASISprob),3)
AUChispanic_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Hispanic",]$OASISprob),3)
AUCwhite_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "White",]$DiedHosp, icustays[icustays$ETHNICITY2 == "White",]$OASISprob),3)

```

The corresponding AUCs (95% CI) are:  
  all people - `r aucAll_M` (`r AUCall_Oasis`),   
Hispanic - `r aucHispanic_M` (`r AUChispanic_Oasis`),  
Black - `r aucBlack_M` (`r AUCblack_Oasis`),  
White - `r aucWhite_M` (`r AUCwhite_Oasis`) and  
Asian - `r aucAsian_M` (`r AUCasian_Oasis`)  

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
# Now test the difference between AUROCs.
rWhite <- roc(response=icustays[icustays$ETHNICITY2 == "White",]$DiedHosp, predictor=icustays[icustays$ETHNICITY2 == "White",]$OASISprob)
rNotWhite <- roc(response=icustays[!(icustays$ETHNICITY2 == "White"),]$DiedHosp, predictor=icustays[!(icustays$ETHNICITY2 == "White"),]$OASISprob)
roc.test(rWhite, rNotWhite)

rBlack <- roc(response=icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp, predictor=icustays[icustays$ETHNICITY2 == "Black",]$OASISprob)
rNotBlack <- roc(response=icustays[!(icustays$ETHNICITY2 == "Black"),]$DiedHosp, predictor=icustays[!(icustays$ETHNICITY2 == "Black"),]$OASISprob)
roc.test(rBlack, rNotBlack)

rAsian <- roc(response=icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp, predictor=icustays[icustays$ETHNICITY2 == "Asian",]$OASISprob)
rNotAsian <- roc(response=icustays[!(icustays$ETHNICITY2 == "Asian"),]$DiedHosp, predictor=icustays[!(icustays$ETHNICITY2 == "Asian"),]$OASISprob)
roc.test(rAsian, rNotAsian)

rHispanic <- roc(response=icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp, predictor=icustays[icustays$ETHNICITY2 == "Hispanic",]$OASISprob)
rNotHispanic <- roc(response=icustays[!(icustays$ETHNICITY2 == "Hispanic"),]$DiedHosp, predictor=icustays[!(icustays$ETHNICITY2 == "Hispanic"),]$OASISprob)
roc.test(rHispanic, rNotHispanic)


```

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, results='hide', fig.width = 12, fig.height=8}
# Now calculate the AUCs for MIMIC-III
sampleSizes_M <- c("n",  as.character(sum(icustays$ETHNICITY2 == "Hispanic")), as.character(sum(icustays$ETHNICITY2 == "Black")), as.character(sum(icustays$ETHNICITY2 == "White")),as.character(sum(icustays$ETHNICITY2 == "Asian")), as.character(nrow(icustays)))

AUCall_Oasis <- round(ci.auc(icustays$DiedHosp, icustays$OASISprob),3)
AUCasian_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Asian",]$OASISprob),3)
AUCblack_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Black",]$OASISprob),3)
AUChispanic_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Hispanic",]$OASISprob),3)
AUCwhite_Oasis <- round(ci.auc(icustays[icustays$ETHNICITY2 == "White",]$DiedHosp, icustays[icustays$ETHNICITY2 == "White",]$OASISprob),3)

plotForest("MIMIC-III",AUChispanic_Oasis, AUCblack_Oasis, AUCwhite_Oasis, AUCasian_Oasis, AUCall_Oasis, "Oasis", "AUC",doPNG=FALSE, sampleSizes_M, tag="Oasis")

```


```{r eval=TRUE, echo=FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=12, fig.height=8}
plotForest("MIMIC-III",round(SMRhispanic_MIMIC_Oa,3), round(SMRblack_MIMIC_Oa,3),  round(SMRwhite_MIMIC_Oa,3),round(SMRasian_MIMIC_Oa,3), round(SMRall_MIMIC_Oa,3), "Oasis score", "SMR",doPNG=FALSE, sampleSizes_M, tag="Oasis")

# restore midline decimal point.
#options(OutDec="\xB7")

```


# SOFA

## Fitting a mortality curve to OASIS and SOFA scores in MIMIC-III.

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, fig.width=figWidth2, fig.height=figHeight2}
plotColours <- c("All"='black',"Hispanic"='red', "Black"='blue', "White"='seagreen4', "Asian"='orange')
logisticPlot(binWidth = 1, database = "MIMIC-III", scoreType = "SOFA", ethnicitiesList, colours=plotColours, ethNames = c(hisp = "Hispanic", black = "Black", white = "White", asian = "Asian"), 'logistic')

SOFAmort <- read.csv("SOFAmort_MIMIC.csv")
icustays$SOFAmort <- SOFAmort[,2]
#----------------------------------------- eICU calibration plot -----------------


#par(bg='white', lwd=2, col = 'black')
calibrate.plot_2(icustays$DiedHosp, icustays$SOFAmort, line.par = list(col='black'), shade.col = adjustcolor('gray', alpha = 0.3), main="Calibration plots for SOFA scores \npredicted mortality in MIMIC")
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "White",]$DiedHosp, icustays[icustays$ETHNICITY2 == "White",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='seagreen4'), shade.col = adjustcolor('seagreen4', alpha = 0.1))
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Asian",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='orange'), shade.col = adjustcolor('orange', alpha = 0.1))
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Black",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='blue'), shade.col = adjustcolor('blue', alpha = 0.1))
calibrate.plot_2(icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Hispanic",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='red'), shade.col = adjustcolor('red', alpha = 0.1))
legend('topleft',legend=c("All","Hispanic","Black","White","Asian"), lwd=c(2,2,2,2,2), col=c('black','red','blue','seagreen4','orange'))

#---------------------------------------------------------------------------------


```


The correlation between the deciles of the Oasis predicted mortality and the observed mortality is very high (`r OASISprobCor`). This suggests that the variable 'oasis_PROB' in the 'oasis table of the mimiciii_derived database is, in fact the directly observed mortality rates by Oasis score.


```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width=figWidth2, fig.height=figHeight2}

deathByOasis_T <- aggregate(icustays[, "DiedHosp"], list(icustays$OASIS), mean)
scoreMax <- max(icustays$OASIS)
  deathByOasis <- data.frame(matrix(nrow = scoreMax+1, ncol = 2))
  colnames(deathByOasis) <- c("score","All")
  deathByOasis$score <- 0:scoreMax
  deathByOasis[,2] <- deathByOasis_T[match(deathByOasis$score, deathByOasis_T[,1]),2]


# Now make sure all possible scores accomodated 0:70
oasisscores <- (min(icustays$OASIS):max(icustays$OASIS))
deathByOasis2 <- data.frame(cbind(oasisscores, rep(0, length(oasisscores))))
deathByOasis2[,2] <- deathByOasis[match(deathByOasis2[,1],deathByOasis[,1]),2]
deathByOasis2[is.na(deathByOasis2[,2]),] <- 0

#O2E_SOFA_MIMIC_All_logistic <- glm(DiedHosp)

plot(x=oasisscores,y=deathByOasis2[,2], main="Plot of mortality by Oasis score in the \nMIMIC-III data", xlab="Oasis score", ylab="Mortality", cex.main=0.9)

```

Now we can examine the relationship of the admission SOFA scores to mortality.

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width=figWidth2, fig.height=figHeight2}
SOFAscores <- min(icustays$SOFA_Adm):max(icustays$SOFA_Adm)
deathBySOFA <- aggregate(icustays$DiedHosp, list(icustays$SOFA_Adm), mean)
plot(SOFAscores, deathBySOFA[,2], main="Mortality by SOFA score in the\nMIMIC-III database.", xlab="SOFA score on admission",ylab="Mortality", cex.main=0.9)
```

\
  
 
```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width=figWidth2, fig.height=figHeight2}
icustays$SOFAcategory <- 0
icustays$SOFAcategory[icustays$SOFA_Adm < 2] <- 1
icustays$SOFAcategory[icustays$SOFA_Adm %in% c(2,3)] <- 2
icustays$SOFAcategory[icustays$SOFA_Adm %in% c(4,5)] <- 3
icustays$SOFAcategory[icustays$SOFA_Adm %in% c(6,7)] <- 4
icustays$SOFAcategory[icustays$SOFA_Adm %in% c(8,9)] <- 5
icustays$SOFAcategory[icustays$SOFA_Adm %in% c(10,11)] <- 6
icustays$SOFAcategory[icustays$SOFA_Adm %in% c(12,13)] <- 7
icustays$SOFAcategory[icustays$SOFA_Adm >= 14] <- 8

deathBySOFAcat <- aggregate(icustays$DiedHosp, list(icustays$SOFAcategory), mean)
deathBySOFAcat_Asian <- aggregate(icustays[icustays$ETHNICITY2 == "Asian", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "Asian", "SOFAcategory"]), mean)
deathBySOFAcat_Black <- aggregate(icustays[icustays$ETHNICITY2 == "Black", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "Black", "SOFAcategory"]), mean)
deathBySOFAcat_White <- aggregate(icustays[icustays$ETHNICITY2 == "White", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "White", "SOFAcategory"]), mean)
deathBySOFAcat_Hispanic <- aggregate(icustays[icustays$ETHNICITY2 == "Hispanic", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "Hispanic", "SOFAcategory"]), mean)
plot(y=deathBySOFAcat[,2], x=1:8, main="Mortality by category of admission SOFA score (MIMIC-III).", xlab="Admission SOFA score",ylab="Mortality", cex.main=0.9)
plot(y=deathBySOFAcat_Asian[,2], x=1:8, main="Mortality in Asian people by category of \nadmission SOFA score.", xlab="Admission SOFA score",ylab="Mortality", cex.main=0.9)
plot(y=deathBySOFAcat_Black[,2], x=1:8, main="Mortality in Black people by category of \nadmission SOFA score (MIMIC-III).", xlab="Admission SOFA score",ylab="Mortality", cex.main=0.9)
plot(y=deathBySOFAcat_White[,2], x=1:8, main="Mortality In white people by category of \nadmission SOFA score (MIMIC-III).", xlab="Admission SOFA score",ylab="Mortality", cex.main=0.9)
plot(y=deathBySOFAcat_Hispanic[,2], x=1:8, main="Mortality Hispanic people by category of \nadmission SOFA score (MIMIC-III).", xlab="Admission SOFA score",ylab="Mortality", cex.main=0.9)


```

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
options(OutDec = ".")
deathBySOFAcatsA <- cbind(as.numeric(round(100*deathBySOFAcat[,2],1)), as.numeric(round(100*deathBySOFAcat_Asian[,2], 1)), as.numeric(round(100*deathBySOFAcat_Black[,2],1)), as.numeric(round(100*deathBySOFAcat_White[,2], 1)), as.numeric(round(100*deathBySOFAcat_Hispanic[,2], 1)))
options(OutDec = ".")   # Needed to prevent errors generated by converting decimals with midline strings to characters in cbind.
deathBySOFAcats <- cbind(c("0,1","2,3","4,5","6,7","8,9","10,11","12,13","14+"), as.matrix(deathBySOFAcatsA))
colnames(deathBySOFAcats) <- c("SOFA category", "All", "Asian", "Black", "White", "Hispanic")

SMRbySOFAcats <- data.frame(cbind(deathBySOFAcats[,1], matrix(round(as.numeric(deathBySOFAcats[,2:6])/as.numeric(deathBySOFAcats[,2]),2),8,5)))
colnames(SMRbySOFAcats) <- c("SOFA category", "All", "Asian", "Black", "White", "Hispanic")
options(OutDec="\xB7")  # Put back the midline dps.

knitr::kable(deathBySOFAcats, col.names = c("SOFA category","All", "Asian", "Black", "White", "Hispanic"), format.args = c(width = 50), caption = "Mortality (%) by SOFA category and ethnic group (MIMIC-III).", align=rep('c', 6)) 
```

\pagebreak

   
```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(SMRbySOFAcats, col.names = c("SOFA category","All", "Asian", "Black", "White", "Hispanic"), format.args = c(width = 50), caption = "SMR by SOFA category and ethnic group (MIMIC-III).", align=rep('c', 6)) 
```

\

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width=figWidth2, fig.height=figHeight2}
options(OutDec = ".")

icustays$SOFAcategory <- 0
icustays$SOFAcategory[icustays$SOFA_Adm < 8] <- 1
icustays$SOFAcategory[icustays$SOFA_Adm %in% 8:11] <- 2
icustays$SOFAcategory[icustays$SOFA_Adm > 11] <- 3

deathBySOFAcat <- aggregate(icustays$DiedHosp, list(icustays$SOFAcategory), mean)
deathBySOFAcat_Asian <- aggregate(icustays[icustays$ETHNICITY2 == "Asian", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "Asian", "SOFAcategory"]), mean)
deathBySOFAcat_Black <- aggregate(icustays[icustays$ETHNICITY2 == "Black", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "Black", "SOFAcategory"]), mean)
deathBySOFAcat_White <- aggregate(icustays[icustays$ETHNICITY2 == "White", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "White", "SOFAcategory"]), mean)
deathBySOFAcat_Hispanic <- aggregate(icustays[icustays$ETHNICITY2 == "Hispanic", "DiedHosp"], list(icustays[icustays$ETHNICITY2 == "Hispanic", "SOFAcategory"]), mean)

theNumbers <- c(nrow(icustays[icustays$SOFAcategory == 1,]), 
                nrow(icustays[icustays$SOFAcategory == 2,]), 
                nrow(icustays[icustays$SOFAcategory == 3,])
              )

deathBySOFAcats2 <- cbind(c("0-7","8-11","11+"), theNumbers, round(100*deathBySOFAcat[,2],1), round(100*deathBySOFAcat_Hispanic[,2], 1), round(100*deathBySOFAcat_Black[,2],1), round(100*deathBySOFAcat_White[,2], 1), round(100*deathBySOFAcat_Asian[,2], 1))
colnames(deathBySOFAcats2) <- c("SOFA category", "Num", "All", "Asian", "Black", "White", "Hispanic")

SMRbySOFAcats2 <- data.frame( cbind( deathBySOFAcats2[,1], theNumbers, matrix( round( 
  as.numeric( deathBySOFAcats2[,3:7]) / as.numeric(deathBySOFAcats2[,3]), 2), 3, 5)))
colnames(SMRbySOFAcats2) <- c("SOFA category", "Num", "All", "Hispanic", "Black", "White", "Asian")

options(OutDec = "\xB7")

knitr::kable(deathBySOFAcats2, col.names = c("SOFA category","Nums","All",  "Hispanic", "Black", "White", "Asian"), format.args = c(width = 50), caption = "Mortality (%) by SOFA category and ethnic group (3 categories) (MIMIC-III).", align=rep('c', 7)) 

```
 
\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(SMRbySOFAcats2, col.names = c("SOFA category","Num", "All", "Hispanic", "Black", "White", "Asian"), format.args = c(width = 50), caption = "SMR by SOFA category and ethnic group (3 categories) (MIMIC-III).", align=rep('c', 6)) 

```


```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, results='hide', fig.width=figWidth2, fig.height=figHeight2}
# Now the ROC for SOFA scores in MIMIC-IIIU
library(ROCR)
thePredictionsAll_SM <- icustays$SOFA_Adm
thePredictionsAsian_SM <- icustays[icustays$ETHNICITY2 == "Asian",]$SOFA_Adm
thePredictionsBlack_SM <- icustays[icustays$ETHNICITY2 == "Black",]$SOFA_Adm
thePredictionsHispanic_SM <- icustays[icustays$ETHNICITY2 == "Hispanic",]$SOFA_Adm
thePredictionsWhite_SM <- icustays[icustays$ETHNICITY2 == "White",]$SOFA_Adm

theLabelsAll_SM <- icustays$DiedHosp
theLabelsAsian_SM <- icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp
theLabelsBlack_SM <- icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp
theLabelsHispanic_SM <- icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp
theLabelsWhite_SM <- icustays[icustays$ETHNICITY2 == "White",]$DiedHosp

predObjAll_SM <- prediction(predictions = thePredictionsAll_SM, labels = theLabelsAll_SM)
aucAll_SM <- round(performance(predObjAll_SM, measure = "auc")@y.values[[1]],3)
perfObjAll_SM <- performance(predObjAll_SM, "tpr","fpr")
predObjAsian_SM <- prediction(predictions = thePredictionsAsian_SM, labels = theLabelsAsian_SM)
aucAsian_SM <- round(performance(predObjAsian_SM, measure = "auc")@y.values[[1]],3)
perfObjAsian_SM <- performance(predObjAsian_SM, "tpr","fpr")
predObjBlack_SM <- prediction(predictions = thePredictionsBlack_SM, labels = theLabelsBlack_SM)
aucBlack_SM <- round(performance(predObjBlack_SM, measure = "auc")@y.values[[1]],3)
perfObjBlack_SM <- performance(predObjBlack_SM, "tpr","fpr")
predObjHispanic_SM <- prediction(predictions = thePredictionsHispanic_SM, labels = theLabelsHispanic_SM)
aucHispanic_SM <- round(performance(predObjHispanic_SM, measure = "auc")@y.values[[1]],3)
perfObjHispanic_SM <- performance(predObjHispanic_SM, "tpr","fpr")
predObjWhite_SM <- prediction(predictions = thePredictionsWhite_SM, labels = theLabelsWhite_SM)
aucWhite_SM <- round(performance(predObjWhite_SM, measure = "auc")@y.values[[1]],3)
perfObjWhite_SM <- performance(predObjWhite_SM, "tpr","fpr")

{
  plot(perfObjAll_SM, main = "ROC curves for the SOFA scores in the MIMIC-III \ndatabase by ethnicity.", lwd=3, cex.main=0.9)
  plot(perfObjAsian_SM, add = TRUE, col = 'orange', lwd=3)
  plot(perfObjBlack_SM, add = TRUE, col = 'blue', lwd=3)
  plot(perfObjHispanic_SM, add = TRUE, col = 'red', lwd=3)
  plot(perfObjWhite_SM, add = TRUE, col = 'seagreen4', lwd=3)
  abline(0,1,col='black', lty=2)
  op <- par(cex = 0.8)
  legend('bottomright',legend = c(paste("All", aucAll_SM),paste("Hispanic",aucHispanic_SM),paste("Black", aucBlack_SM),  paste("White",aucWhite_SM), paste("Asian", aucAsian_SM)), col=c("black","red","blue","seagreen4","orange"), lty = c(1,1,1,1,1), lwd=c(3,3,3,3,3))
}

```
 

 
 
```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
LR_SOFA_MIMIC_Eth <- glm(DiedHosp ~ ETHNICITY2 + SOFA_Adm, data = icustays,family=binomial(link=logit))
R2_SOFA_MIMIC_Eth <- rsq(LR_SOFA_MIMIC_Eth)

knitr::kable(round(summary(LR_SOFA_MIMIC_Eth)$coefficients,3), format.args = c(width=70), caption = "Logistic regression of ethnicity and in-hospital death controlled for SOFA score (MIMIC-II).", align='c')

```
 
 
\pagebreak
# eICU analysis

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
# Now or eICU data

# This data is from the eICU database on www.physionet.org
# This is the 'Patient' table.
eICU_Pats <-  read.csv("bq-results-20200809-141932-y1zn72dtxh4e_EICU_PATIENT.csv")
Pats_TotNo <- nrow(eICU_Pats)
# This is the 'apachepatientresult' table.
eICU_ApacheResFull <- read.csv("bq-results-20200809-142217-bcushyutuc8s_EICU_ApacheRes.csv")
# This is a table of SOFA scores derived from the eICU database. See publication for details.
eICU_SOFA <- read.csv("bq-results-20200810-120702-s3rw32q3hplb-eICU_SOFA.csv")


# Only look at Apache IVa entries.
notApache4 <- nrow(eICU_ApacheResFull[!eICU_ApacheResFull$apacheversion == "IVa",])
eICU_ApacheResFull <- eICU_ApacheResFull[eICU_ApacheResFull$apacheversion == "IVa",] 

# remember the original total of episodes.
ApacheRes_N_Original <- nrow(eICU_ApacheResFull)

# Fetch the ethnicity classifications and put them into the working table.
eICU_ApacheResFull$ethnicity <- eICU_Pats$ethnicity[match(eICU_ApacheResFull$patientunitstayid, eICU_Pats$patientunitstayid)]
eICU_ApacheResFull[eICU_ApacheResFull$ethnicity == "Caucasian",]$ethnicity <- "White"
eICU_ApacheResFull[eICU_ApacheResFull$ethnicity == "African American",]$ethnicity <- "Black"

# Count the episodes out of the age range.
# Calculate in the age.
eICU_ApacheResFull$age <- eICU_Pats$age[match(eICU_ApacheResFull$patientunitstayid, eICU_Pats$patientunitstayid)]

# Copy in the gender
eICU_ApacheResFull$gender <- eICU_Pats$gender[match(eICU_ApacheResFull$patientunitstayid, eICU_Pats$patientunitstayid)]

# Count the episodes in ages out of scope
eICU_under16 <- nrow(eICU_ApacheResFull[eICU_ApacheResFull$age %in% 0:15,])
eICU_over90 <- nrow(eICU_ApacheResFull[eICU_ApacheResFull$age == "> 89",])
# Count the missing age entries
eicu_ageBlank <- nrow(eICU_ApacheResFull[(eICU_ApacheResFull$age == ""),])

# Now drop the out of scope or missing ages.
eICU_ApacheResFull <- eICU_ApacheResFull[!(eICU_ApacheResFull$age %in% 0:15),]
eICU_ApacheResFull <- eICU_ApacheResFull[!(eICU_ApacheResFull$age == "> 89"),]
eICU_ApacheResFull <- eICU_ApacheResFull[!(eICU_ApacheResFull$age == ""),]

# Remember the number of episodes in our age range
eICU_epNum <- nrow(eICU_ApacheResFull)

# Do counts of dropped ethnicity entries.
eICU_ethMiss1 <- nrow(eICU_ApacheResFull[eICU_ApacheResFull$ethnicity == "",])
eICU_ethMiss2 <- nrow(eICU_ApacheResFull[is.na(eICU_ApacheResFull$ethnicity),])
eICU_ethUnknown <- nrow(eICU_ApacheResFull[(eICU_ApacheResFull$ethnicity == "Other/Unknown"),])
eICU_ethMiss <- eICU_ethMiss1 + eICU_ethMiss2 + eICU_ethUnknown
eICU_NatAm <- nrow(eICU_ApacheResFull[eICU_ApacheResFull$ethnicity == "Native American",])
eICU_ethDropped <- eICU_ethMiss + eICU_NatAm 

# Count the episodes with missing Apache scores or Apache predicted mortalities.
eICU_mortMiss <- nrow(eICU_ApacheResFull[!(eICU_ApacheResFull$predictedhospitalmortality >= 0), ])
eICU_noScore <-  nrow(eICU_ApacheResFull[!(eICU_ApacheResFull$apachescore >= 0),])

# Drop APache & Mort <0
eICU_ApacheResFull <- eICU_ApacheResFull[(eICU_ApacheResFull$apachescore >= 0) & (eICU_ApacheResFull$predictedhospitalmortality >= 0), ]

# Drop those admissions with no ethnicity recorded
eICU_ApacheResFull <- eICU_ApacheResFull[!is.na(eICU_ApacheResFull$ethnicity),]   # Remove the NAs
eICU_ApacheResFull <- eICU_ApacheResFull[!(eICU_ApacheResFull$ethnicity == ""),]  # Remove the missing
eICU_ApacheResFull <- eICU_ApacheResFull[!(eICU_ApacheResFull$ethnicity == "Other/Unknown"),] # Remove the 'other/unknown'
eICU_ApacheResFull <- eICU_ApacheResFull[!(eICU_ApacheResFull$ethnicity == "Native American"),] # Remove the Native Americans


```

\
# Data cleaning
We restricted our analysis to Apache IVa scores, and there were `r nrow(ApacheRes_N_Original)` entries altogether.
After removing the ages out of scope there were `r eICU_epNum` episodes included.

There were `r eICU_ethMiss1` episodes with no recorded ethnicity, making `r round(eICU_ethMiss1/eICU_epNum*100,1)`% of the total.

There were `r eICU_ethMiss2` episodes with an ethnicity entry of 'NA', making `r round(eICU_ethMiss2/eICU_epNum*100,1)`% of the total. 

There were `r eICU_ethUnknown` episodes with an ethnicity entry of 'other/unknown', making `r round(eICU_ethUnknown/eICU_epNum*100,1)`% of the total.

There were `r eICU_NatAm` episodes in Native Americans who were also excluded, making `r round(eICU_NatAm/eICU_epNum*100,1)`% of the total.

Altogether there were `r eICU_ethDropped` that were dropped on account of the ethnicity entries, making `r round(eICU_ethDropped/eICU_epNum*100,1)`% of the total.

`r eICU_mortMiss` episodes were dropped as they did not have a valid predicted mortality (`r round(eICU_mortMiss/eICU_epNum*100, 1)`%). 

`r eICU_noScore` episodes did not have an Apache score (`r round(eICU_noScore/eICU_epNum*100, 1)`%). 

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

eICU_Dat <- eICU_ApacheResFull     

# Drop any that are not in our list of ethnicities.
eICU_Dat <- eICU_Dat[(eICU_Dat$ethnicity %in% ethnicitiesList),]

#######################
eICU_Dat$DiedHosp <- FALSE
eICU_Dat$DiedHosp <- eICU_ApacheResFull$actualhospitalmortality[match(eICU_Dat$patientunitstayid, eICU_ApacheResFull$patientunitstayid)]
eICU_Dat$DiedHosp[eICU_Dat$DiedHosp == "ALIVE"] <- 0
eICU_Dat$DiedHosp[eICU_Dat$DiedHosp == "EXPIRED"] <- 1
eICU_Dat$DiedHosp <- as.numeric(eICU_Dat$DiedHosp)
eICU_deathMiss <- nrow(eICU_Dat[is.na(eICU_Dat$DiedHosp),])
eICU_Dat <- eICU_Dat[!is.na(eICU_Dat$DiedHosp),]
# ICU
eICU_Dat$DiedICU <- FALSE
eICU_Dat$DiedICU <- eICU_ApacheResFull$actualicumortality[match(eICU_Dat$patientunitstayid, eICU_ApacheResFull$patientunitstayid)]
eICU_Dat$DiedICU[eICU_Dat$DiedICU == "ALIVE"] <- 0
eICU_Dat$DiedICU[eICU_Dat$DiedICU == "EXPIRED"] <- 1
eICU_Dat$DiedICU <- as.numeric(eICU_Dat$DiedICU)
eICU_Dat <- eICU_Dat[!is.na(eICU_Dat$DiedICU),]

# Add the Apache IVa scores
eICU_Dat$apache <- eICU_ApacheResFull$apachescore[match(eICU_Dat$patientunitstayid, eICU_ApacheResFull$patientunitstayid)]
eICU_Dat$apacheMort <- eICU_ApacheResFull$predictedhospitalmortality[match(eICU_Dat$patientunitstayid, eICU_ApacheResFull$patientunitstayid)]
# Drop the '-1' mortality rates
missingApacheMort <- nrow(eICU_Dat[eICU_Dat$apacheMort == 0,])
eICU_Dat <- eICU_Dat[eICU_Dat$apacheMort >= 0,]

############
eICU_SOFA1 <- eICU_SOFA[eICU_SOFA$day == 1,]
eICU_Dat$SOFA1 <- eICU_SOFA1$sofa[match(eICU_Dat$patientunitstayid, eICU_SOFA1$patientunitstayid)]

# Drop any rows with missing SOFA scores.
eICU_Dat <- eICU_Dat[!is.na(eICU_Dat$SOFA1),] 

```



  
```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

eICU_Dat$SOFAcategory <- 0
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 < 2] <- 1
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 %in% c(2,3)] <- 2
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 %in% c(4,5)] <- 3
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 %in% c(6,7)] <- 4
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 %in% c(8,9)] <- 5
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 %in% c(10,11)] <- 6
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 %in% c(12,13)] <- 7
eICU_Dat$SOFAcategory[eICU_Dat$SOFA1 >= 14] <- 8

eICU_Dat[,c("age")] <- as.numeric(eICU_Dat[,c("age")])
eICU_Dat$gender[eICU_Dat$gender == "Female"] <- 1      # 1 = Female
eICU_Dat$gender[eICU_Dat$gender == "Male"] <- 0
eICU_Dat$gender <- as.numeric(eICU_Dat$gender)

deathBySOFA <- aggregate(as.numeric(eICU_Dat$DiedHosp), list(eICU_Dat$SOFAcategory), mean)
deathBySOFA[,2] <- round(100*deathBySOFA[,2], 2)
colnames(deathBySOFA) <- c("SOFA category", "Mortality")
deathBySOFA_Asian <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Asian", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "Asian", "SOFAcategory"]), mean)
deathBySOFA_Asian[,2] <- round(100*deathBySOFA_Asian[,2], 2)
colnames(deathBySOFA_Asian) <- c("SOFA category", "Mortality")
deathBySOFA_Black <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Black", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "Black", "SOFAcategory"]), mean)
deathBySOFA_Black[,2] <- round(100*deathBySOFA_Black[,2], 2)
colnames(deathBySOFA_Black) <- c("SOFA category", "Mortality")
deathBySOFA_White <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "White", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "White", "SOFAcategory"]), mean)
deathBySOFA_White[,2] <- round(100*deathBySOFA_White[,2], 2)
colnames(deathBySOFA_White) <- c("SOFA category", "Mortality")
deathBySOFA_Hispanic <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Hispanic", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "Hispanic", "SOFAcategory"]), mean)
deathBySOFA_Hispanic[,2] <- round(100*deathBySOFA_Hispanic[,2], 2)
colnames(deathBySOFA_Hispanic) <- c("SOFA category", "Mortality")

options(OutDec = ".")
deathBySOFAsum <- cbind(c("0,1","2,3","4,5","6,7","8,9","10,11","12,13","14+"), deathBySOFA[,2], deathBySOFA_Asian[,2], deathBySOFA_Black[,2], deathBySOFA_White[,2], deathBySOFA_Hispanic[,2])
colnames(deathBySOFAsum) <- c("SOFA category", "All", "Asian", "Black", "White", "Hispanic")

SMRbySOFA_sum <- data.frame(cbind(deathBySOFA[,1], matrix(round(as.numeric(deathBySOFAsum[,2:6])/as.numeric(deathBySOFAsum[,2]),2),8,5)))
colnames(SMRbySOFA_sum) <- c("SOFA category", "All", "Asian", "Black", "White", "Hispanic")
options(OutDec = "\xB7")
knitr::kable(table(eICU_Dat$ethnicity), format.args = c(width=70), caption = "Number of ITU episodes in the target ethnic groups once episodes with missing data are removed (eICU).", align='c')
```
 
   
 

```{r eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}

theTable <- aggregate(eICU_Dat[,c("age", "gender", "apachescore", "predictedhospitalmortality", "DiedHosp", "DiedICU")], by=list(eICU_Dat$ethnicity), FUN = function(x){ paste0(mean=round(mean(x, na.rm = TRUE),2), " (", sd = round(sd(x, na.rm = TRUE),2), ")")})
colnames(theTable) <- c("Ethnicity", "age", "proportion female", "apache score", "Predicted mortality", "Actual mortality", "Death in ICU")

knitr::kable(theTable, caption = "Baseline characteristics for the eICU data.")

```
  
  

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

# First table ===========
baseTable1 <- data.frame(matrix(ncol=11, nrow=0))
for(eth in ethnicitiesList){
  eICU_DatX <- eICU_Dat[eICU_Dat$ethnicity == eth,]
  NewRow <- as.matrix(t(c(nrow(eICU_DatX), eth, round(median(eICU_DatX$age, na.rm=TRUE),2), paste(round(quantile(eICU_DatX$age, 0.25, na.rm=TRUE),2), "-",round(quantile(eICU_DatX$age, 0.75, na.rm=TRUE),1)), round(mean(eICU_DatX$gender, na.rm=TRUE),2), median(eICU_DatX$apachescore, na.rm=TRUE), paste(quantile(eICU_DatX$apachescore, 0.25), "-",quantile(eICU_DatX$apachescore, 0.75)),round(mean(eICU_DatX$predictedhospitalmortality, na.rm=TRUE),4), round(sd(eICU_DatX$predictedhospitalmortality, na.rm=TRUE),4), round(mean(eICU_DatX$DiedHosp, na.rm=TRUE),3), round(mean(eICU_DatX$DiedICU, na.rm=TRUE),3))))   
  baseTable1 <- rbind(baseTable1, NewRow)
}
NewRow <- as.matrix(t(c(nrow(eICU_Dat), "All", round(median(eICU_Dat$age, na.rm=TRUE),1), paste(round(quantile(eICU_Dat$age, 0.25, na.rm=TRUE),1), "-",round(quantile(eICU_Dat$age, 0.75, na.rm=TRUE),1)), round(mean(eICU_Dat$gender, na.rm=TRUE),2), median(eICU_Dat$apachescore, na.rm=TRUE), paste(quantile(eICU_Dat$apachescore, 0.25), "-",quantile(eICU_Dat$apachescore, 0.75)),round(mean(eICU_Dat$predictedhospitalmortality, na.rm=TRUE),4), round(sd(eICU_Dat$predictedhospitalmortality, na.rm=TRUE),4), round(mean(eICU_Dat$DiedHosp, na.rm=TRUE),3), round(mean(eICU_Dat$DiedICU, na.rm=TRUE),3)))) 
baseTable1 <- rbind(baseTable1, NewRow)
colnames(baseTable1) <- c("Num", "Ethnicity", "median age","IQR age", "proportion female","median Apache score", "IQR Apache score", "mean predicted hospital mortality", "sd predicted hospital mortality", "hospital mortality", "ICU mortality")
rownames(baseTable1) <- c(ethnicitiesList, "All")

# Second table ===========
baseTable2 <- data.frame(matrix(ncol=6, nrow=0))
for(eth in ethnicitiesList){
  eICU_DatX <- eICU_Dat[eICU_Dat$ethnicity == eth,]
  NewRow <- as.matrix(t(c(round(median(eICU_DatX$unabridgedunitlos, na.rm=TRUE),1), paste(round(quantile(eICU_DatX$unabridgedunitlos, 0.25),1), "-",round(quantile(eICU_DatX$unabridgedunitlos, 0.75),1)), round(median(eICU_DatX$unabridgedhosplos, na.rm=TRUE),1), paste(round(quantile(eICU_DatX$unabridgedhosplos, 0.25),1), "-", round(quantile(eICU_DatX$unabridgedhosplos, 0.75),1)))))   
  baseTable2 <- rbind(baseTable2, NewRow)
}
NewRow <- as.matrix(t(c(round(median(eICU_Dat$unabridgedunitlos, na.rm=TRUE),1), paste(round(quantile(eICU_Dat$unabridgedunitlos, 0.25),1), "-",round(quantile(eICU_Dat$unabridgedunitlos, 0.75),1)), round(median(eICU_Dat$unabridgedhosplos, na.rm=TRUE),1), paste(round(quantile(eICU_Dat$unabridgedhosplos, 0.25),1), "-", round(quantile(eICU_Dat$unabridgedhosplos, 0.75),1)))))   
  baseTable2 <- rbind(baseTable2, NewRow)
rownames(baseTable2) <- c(ethnicitiesList, "All")
colnames(baseTable2) <- c("median ICU LoS", "IQR ICU LoS", "median hospital LOS", "IQR Hospital LoS")

knitr::kable(baseTable1,format.args=c(width=100), caption = "Baseline characteristics for eICU patients.", align='c')
```

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(baseTable2,format.args=c(width=100), caption = "Baseline characteristics for eICU patients cont'd.", align='c')

```

 
# Apache

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, fig.width=figWidth2, fig.height=figHeight2}
plotColours <- c("All"='black',"Hispanic"='red', "Black"='blue', "White"='seagreen4', "Asian"='orange')
logisticPlot(binWidth = 1, database = "eICU", scoreType = "Apache", ethnicitiesList, colours=plotColours, ethNames = c(hisp = "Hispanic", black = "Black", white = "White", asian = "Asian"))
```

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

#par(bg='white', lwd=2, col = 'black')
calibrate.plot_2(eICU_Dat$DiedHosp, eICU_Dat$predictedhospitalmortality, line.par = list(col='black'), shade.col = adjustcolor('gray', alpha = 0.1), main="Calibration plots for APACHE IVa in eICU")
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "White",]$predictedhospitalmortality, replace = FALSE, line.par = list(lwd=2, col='seagreen4'), shade.col = adjustcolor('seagreen4', alpha = 0.1))
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Asian",]$predictedhospitalmortality, replace = FALSE, line.par = list(lwd=2, col='orange'), shade.col = adjustcolor('orange', alpha = 0.1))
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Black",]$predictedhospitalmortality, replace = FALSE, line.par = list(lwd=2, col='blue'), shade.col = adjustcolor('blue', alpha = 0.1))
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$predictedhospitalmortality, replace = FALSE, line.par = list(lwd=2, col='red'), shade.col = adjustcolor('red', alpha = 0.1))
legend('topleft',legend=c("All","Hispanic","Black","White","Asian"), lwd=c(2,2,2,2,2), col=c('black','red','blue','seagreen4','orange'))

```


This figure shows the ROC curves for Apache IVa predicted mortality in the four principle ethnic groups.

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=figWidth2, fig.height=figHeight2}
library(ROCR)
thePredictionsAll <- eICU_Dat$apacheMort
thePredictionsAsian <- eICU_Dat[eICU_Dat$ethnicity == "Asian",]$apacheMort
thePredictionsBlack <- eICU_Dat[eICU_Dat$ethnicity == "Black",]$apacheMort
thePredictionsHispanic <- eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$apacheMort
thePredictionsWhite <- eICU_Dat[eICU_Dat$ethnicity == "White",]$apacheMort

theLabelsAll <- eICU_Dat$DiedHosp
theLabelsAsian <- eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp
theLabelsBlack <- eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp
theLabelsHispanic <- eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp
theLabelsWhite <- eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp

predObjAll <- prediction(predictions = thePredictionsAll, labels = theLabelsAll)
aucAll <- round(performance(predObjAll, measure = "auc")@y.values[[1]],3)
perfObjAll <- performance(predObjAll, "tpr","fpr")
predObjAsian <- prediction(predictions = thePredictionsAsian, labels = theLabelsAsian)
aucAsian <- round(performance(predObjAsian, measure = "auc")@y.values[[1]],3)
perfObjAsian <- performance(predObjAsian, "tpr","fpr")
predObjBlack <- prediction(predictions = thePredictionsBlack, labels = theLabelsBlack)
aucBlack <- round(performance(predObjBlack, measure = "auc")@y.values[[1]],3)
perfObjBlack <- performance(predObjBlack, "tpr","fpr")
predObjHispanic <- prediction(predictions = thePredictionsHispanic, labels = theLabelsHispanic)
aucHispanic <- round(performance(predObjHispanic, measure = "auc")@y.values[[1]],3)
perfObjHispanic <- performance(predObjHispanic, "tpr","fpr")
predObjWhite <- prediction(predictions = thePredictionsWhite, labels = theLabelsWhite)
aucWhite <- round(performance(predObjWhite, measure = "auc")@y.values[[1]],3)
perfObjWhite <- performance(predObjWhite, "tpr","fpr")

{
  plot(perfObjAll, main = "ROC curves for the Apache IVa predicted hospital \nmortality in the eICU database by ethnicity.", lwd=3, cex.main=0.9)
  plot(perfObjAsian, add = TRUE, col = 'orange', lwd=3)
  plot(perfObjBlack, add = TRUE, col = 'blue', lwd=3)
  plot(perfObjHispanic, add = TRUE, col = 'red', lwd=3)
  plot(perfObjWhite, add = TRUE, col = 'seagreen4', lwd=3)
  abline(0,1,col='black', lty=2)
  op <- par(cex = 0.8)
  legend('bottomright',legend = c(paste("All", aucAll),paste("Hispanic",aucHispanic),paste("Black", aucBlack), paste("White",aucWhite),paste("Asian", aucAsian)), col=c("black","red","blue","seagreen4","orange"), lty = c(1,1,1,1,1), lwd=c(3,3,3,3,3))
}

```


 
\pagebreak

# Now determine the p value for ethnicity as a predictor of each of the baseline variables.

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

pAge <- lm(eICU_Dat$age ~ eICU_Dat$ethnicity)
print(summary(pAge))
pgender <- lm(eICU_Dat$gender ~ eICU_Dat$ethnicity)
print(summary(pgender))
```

\pagebreak
 
```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
pLoS <- lm(eICU_Dat$unabridgedunitlos ~ eICU_Dat$ethnicity)
print(summary(pLoS))
pDiedHosp <- lm(eICU_Dat$DiedHosp ~ eICU_Dat$ethnicity)
print(summary(pDiedHosp))
```

\pagebreak
 
```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
pDiedICU <- lm(eICU_Dat$DiedICU ~ eICU_Dat$ethnicity)
print(summary(pDiedICU))
```

\pagebreak


```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
pPredMort <- lm(eICU_Dat$predictedicumortality ~ eICU_Dat$ethnicity)
print(summary(pPredMort))
pICUmort <- lm(eICU_Dat$predictedicumortality ~ eICU_Dat$ethnicity)
print(summary(pICUmort))

```

\pagebreak

 
   
```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

AUCall_Apache <- round(ci.auc(eICU_Dat$DiedHosp, eICU_Dat$apacheMort),3)
AUCasian_Apache <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Asian",]$apacheMort),3)
AUCblack_Apache <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Black",]$apacheMort),3)
AUChispanic_Apache <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$apacheMort),3)
AUCwhite_Apache <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "White",]$apacheMort),3)

```
 
The corresponding AUCs (95% CI) are:  
All - `r aucAll` (`r AUCall_Apache`),  
Hispanic - `r aucHispanic` (`r AUChispanic_Apache`),  
Black - `r aucBlack` (`r AUCblack_Apache`),  
White - `r aucWhite` (`r AUCwhite_Apache`) and  
Asian -  , `r aucAsian` (`r AUCasian_Apache`).
 
\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
# Now test the difference between ROCs.
rAPACHE_eICU <- roc(response = eICU_Dat$DiedHosp, predictor = eICU_Dat$apacheMort)

rWhite_eICU <- roc(response=eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp, predictor = eICU_Dat[eICU_Dat$ethnicity == "White",]$apacheMort)
rNotWhite_eICU <- roc(response=eICU_Dat[!(eICU_Dat$ethnicity == "White"),]$DiedHosp, predictor = eICU_Dat[!(eICU_Dat$ethnicity == "White"),]$apacheMort)
roc.test(rWhite_eICU, rNotWhite_eICU)
roc.test(rWhite_eICU, rAPACHE_eICU)

rBlack_eICU <- roc(response=eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp, predictor = eICU_Dat[eICU_Dat$ethnicity == "Black",]$apacheMort)
rNotBlack_eICU <- roc(response=eICU_Dat[!(eICU_Dat$ethnicity == "Black"),]$DiedHosp, 
                          predictor = eICU_Dat[!(eICU_Dat$ethnicity == "Black"),]$apacheMort)
roc.test(rBlack_eICU, rNotBlack_eICU)
roc.test(rBlack_eICU, rAPACHE_eICU)

rAsian_eICU <- roc(response=eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp, predictor = eICU_Dat[eICU_Dat$ethnicity == "Asian",]$apacheMort)
rNotAsian_eICU <- roc(response=eICU_Dat[!(eICU_Dat$ethnicity == "Asian"),]$DiedHosp, predictor = eICU_Dat[!(eICU_Dat$ethnicity == "Asian"),]$apacheMort)
roc.test(rAsian_eICU, rNotAsian_eICU)
roc.test(rAsian_eICU, rAPACHE_eICU)

rHispanic_eICU <- roc(response=eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp, predictor = eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$apacheMort)
rNotHispanic_eICU <- roc(response=eICU_Dat[!(eICU_Dat$ethnicity == "Hispanic"),]$DiedHosp, predictor = eICU_Dat[!(eICU_Dat$ethnicity == "Hispanic"),]$apacheMort)
roc.test(rHispanic_eICU, rNotHispanic_eICU)
roc.test(rHispanic_eICU, rAPACHE_eICU)

```


```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, results='hide', fig.width = 12, fig.height=8}

sampleSizes_eICU <- c("n",  as.character(sum(eICU_Dat$ethnicity == "Hispanic")), as.character(sum(eICU_Dat$ethnicity == "Black")), as.character(sum(eICU_Dat$ethnicity == "White")),as.character(sum(eICU_Dat$ethnicity == "Asian")), as.character(nrow(eICU_Dat)))
plotForest("eICU",AUChispanic_Apache, AUCblack_Apache, AUCwhite_Apache,AUCasian_Apache, AUCall_Apache, "Apache IVa", "AUC",doPNG=FALSE, sampleSizes_eICU, tag="Apache")
```


```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

# eICU - Apache
SMRall_eICU_Ap <- SMR(as.numeric(eICU_Dat$DiedHosp), eICU_Dat$apacheMort)
SMRall_eICU_Ap <- c(SMRall_eICU_Ap[5], SMRall_eICU_Ap[4], SMRall_eICU_Ap[6])
SMRasian_eICU_Ap <- SMR(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp), eICU_Dat[eICU_Dat$ethnicity == "Asian",]$apacheMort)
SMRasian_eICU_Ap <- c(SMRasian_eICU_Ap[5], SMRasian_eICU_Ap[4], SMRasian_eICU_Ap[6])
SMRblack_eICU_Ap <- SMR(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp), eICU_Dat[eICU_Dat$ethnicity == "Black",]$apacheMort)
SMRblack_eICU_Ap <- c(SMRblack_eICU_Ap[5], SMRblack_eICU_Ap[4], SMRblack_eICU_Ap[6])
SMRhispanic_eICU_Ap <- SMR(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp), eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$apacheMort)
SMRhispanic_eICU_Ap <- c(SMRhispanic_eICU_Ap[5], SMRhispanic_eICU_Ap[4], SMRhispanic_eICU_Ap[6])
SMRwhite_eICU_Ap <- SMR(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp), eICU_Dat[eICU_Dat$ethnicity == "White",]$apacheMort)
SMRwhite_eICU_Ap <- c(SMRwhite_eICU_Ap[5], SMRwhite_eICU_Ap[4], SMRwhite_eICU_Ap[6])

```

```{r eval = TRUE, echo=FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=12, fig.height=8}
sampleSizes_eICU <- c("n",  as.character(sum(eICU_Dat$ethnicity == "Hispanic")), as.character(sum(eICU_Dat$ethnicity == "Black")))
plotForest("eICU",round(SMRhispanic_eICU_Ap,3), round(SMRblack_eICU_Ap,3), round(SMRwhite_eICU_Ap,3), round(SMRasian_eICU_Ap,3), round(SMRall_eICU_Ap,3), "Apache score", "SMR",doPNG=FALSE, sampleSizes_eICU, tag="Apache")

```

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}


total_eICU <- eICU_Dat[,c("age", "ethnicity", "predictedhospitalmortality", "DiedHosp", "SOFA1", "SOFAcategory")]
colnames(total_eICU) <- c("AGE", "ETHNICITY2", "PREDICTED", "DiedHosp", "SOFA", "SOFAcategory")
total_eICU[total_eICU$ETHNICITY2 == "White","ETHNICITY2"] <- "White"
total_eICU[total_eICU$ETHNICITY2 == "African American","ETHNICITY2"] <- "Black"
total_eICU[total_eICU$ETHNICITY2 == "Asian","ETHNICITY2"] <- "Asian"
total_eICU[total_eICU$ETHNICITY2 == "Hispanic","ETHNICITY2"] <- "Hispanic"

total_MIMIC <- icustays[, c("AGE", "ETHNICITY2", "OASISprob", "DiedHosp", "SOFA_Adm", "SOFAcategory")]
colnames(total_MIMIC) <- c("AGE", "ETHNICITY2", "PREDICTED", "DiedHosp", "SOFA", "SOFAcategory")
total_MIMIC[total_MIMIC$DiedHosp == "TRUE", "DiedHosp"] <- 1
total_MIMIC[total_MIMIC$DiedHosp == "FALSE", "DiedHosp"] <- 0
total_eICU[total_eICU$DiedHosp == "TRUE", "DiedHosp"] <- 1
total_eICU[total_eICU$DiedHosp == "FALSE", "DiedHosp"] <- 0
total_MIMIC$AGE <- round(total_MIMIC$AGE, 0)

cutPoints <- c(0, 0.05, 0.1, 0.2, 0.5, 1.0)

MIMIC_SMRbyRiskCat <- matrix(rep(0, length(cutPoints)-1)*length(ethnicitiesList), length(ethnicitiesList), length(cutPoints)-1)
rownames(MIMIC_SMRbyRiskCat) <- ethnicitiesList
colnames(MIMIC_SMRbyRiskCat) <- c(" 0-5%"," 5-10%","10-20%", "20-50%","50-100%" )

MIMIC_byRiskCat <- list()
catNames <- c(" 0-5%"," 5-10%","10-20%", "20-50%","50-100%" )
icustays$riskSeg <- NA
for(i in 2:length(cutPoints)){
  icustays[icustays$OASISprob >= cutPoints[i-1] & icustays$OASISprob < cutPoints[i],]$riskSeg <- catNames[i-1]
}

MIMIC_byRiskEthn <- aggregate(icustays[,c("OASISprob", "DiedHosp")], list(icustays$ETHNICITY2, icustays$riskSeg), mean)
MIMIC_byRiskEthn$SMR <- round(MIMIC_byRiskEthn[,4]/MIMIC_byRiskEthn[,3],2)
colnames(MIMIC_byRiskEthn) <- c("Ethnicity", "riskGroup", "Risk", "Mort", "SMR")

MIMIC_SMRbyRiskTable <- dcast(MIMIC_byRiskEthn[,c("Ethnicity","riskGroup","SMR")], Ethnicity ~ riskGroup)

knitr::kable(MIMIC_SMRbyRiskTable, align=c('r','c'), caption="SMRs segmented by ethnicity and risk group in MIMIC-III.")

```


```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}


total_predMort <- rbind(as.matrix(eICU_Dat[,c("age", "ethnicity", "predictedhospitalmortality", "DiedHosp")]), as.matrix(icustays[, c("AGE", "ETHNICITY2", "OASISprob", "DiedHosp")]))
icustays$riskSeg <- NA
eICU_Dat$riskSeg <- NA

cutPoints <- c(0, 0.05, 0.1, 0.2, 0.5, 1.0)
eICU_SMRbyRiskCat <- data.frame(matrix(rep(0, length(cutPoints)-1)*length(ethnicitiesList), length(ethnicitiesList), length(cutPoints)-1))
rownames(eICU_SMRbyRiskCat) <- ethnicitiesList
colnames(eICU_SMRbyRiskCat) <- c("0-5%","5-10%","10-20%", "20-50%","50-100%" )

eICU_byRiskCat <- list()
catNames <- c(" 0-5%"," 5-10%","10-20%", "20-50%","50-100%" )
for(i in 2:length(cutPoints)){
	eICU_Dat[eICU_Dat$predictedhospitalmortality >= cutPoints[i-1] & eICU_Dat$predictedhospitalmortality < cutPoints[i],]$riskSeg <- catNames[i-1]
}
eiCU_byRiskEthn <- aggregate(eICU_Dat[,c("predictedhospitalmortality", "DiedHosp")], list(eICU_Dat$ethnicity, eICU_Dat$riskSeg), mean)
eiCU_byRiskEthn$SMR <- round(eiCU_byRiskEthn[,4]/eiCU_byRiskEthn[,3],2)
colnames(eiCU_byRiskEthn) <- c("Ethnicity", "riskGroup", "Risk", "Mort", "SMR")

eICU_SMRbyRiskTable <- dcast(eiCU_byRiskEthn[,c("Ethnicity","riskGroup","SMR")], Ethnicity ~ riskGroup)
knitr::kable(eICU_SMRbyRiskTable, align=c('r','c'), caption="SMRs segmented by ethnicity and risk group in eICU.")

```


## SOFA analysis

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE, fig.width=figWidth2, fig.height=figHeight2}
plotColours <- c("All"='black',"Hispanic"='red', "Black"='blue', "White"='seagreen4', "Asian"='orange')
logisticPlot(binWidth = 1, database = "eICU", scoreType = "SOFA", ethnicitiesList, colours=plotColours, ethNames = c(hisp = "Hispanic", black = "Black", white = "White", asian = "Asian"))
SOFAmort <- read.csv("SOFAmort_eICU.csv")
eICU_Dat$SOFAmort <- SOFAmort[,2]
#----------------------------------------- eICU calibration plot -----------------


#par(bg='white', lwd=2, col = 'black')
calibrate.plot_2(eICU_Dat$DiedHosp, eICU_Dat$SOFAmort, line.par = list(col='black'), shade.col = adjustcolor('gray', alpha = 0.3), main="Calibration plots for SOFA scores \npredicted mortality in eICU")
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "White",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='seagreen4'), shade.col = adjustcolor('seagreen4', alpha = 0.1))
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Asian",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='orange'), shade.col = adjustcolor('orange', alpha = 0.1))
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Black",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='blue'), shade.col = adjustcolor('blue', alpha = 0.1))
calibrate.plot_2(eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$SOFAmort, replace = FALSE, line.par = list(lwd=2, col='red'), shade.col = adjustcolor('red', alpha = 0.1))
legend('topleft',legend=c("All","Hispanic","Black","White","Asian"), lwd=c(2,2,2,2,2), col=c('black','red','blue','seagreen4','orange'))

#---------------------------------------------------------------------------------
  
```



```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(deathBySOFAsum, format.args = c(width=70), caption = "Mortality rates by ethnic group and day 1 SOFA category (eICU).", align='c')

```
 
\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(SMRbySOFA_sum, format.args = c(width=70), caption = "Mortality in the eICU database by ethnic group and day 1 SOFA category standardised to the total mortality for the SOFA category (eICU).", align='c')
```

\

The total number of episodes upon which the SOFA analysis in the eICU database is based is `r nrow(eICU_Dat)`.


```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
options(OutDec = ".")
eICU_Dat$SOFAcategory2 <- 0
eICU_Dat$SOFAcategory2[eICU_Dat$SOFA1 < 8] <- 1
eICU_Dat$SOFAcategory2[eICU_Dat$SOFA1 %in% 8:11] <- 2
eICU_Dat$SOFAcategory2[eICU_Dat$SOFA1 > 11] <- 3

eICU_Dat[,c("age")] <- as.numeric(eICU_Dat[,c("age")])
eICU_Dat$gender[eICU_Dat$gender == "Female"] <- 1      # 1 = Female
eICU_Dat$gender[eICU_Dat$gender == "Male"] <- 0
eICU_Dat$gender <- as.numeric(eICU_Dat$gender)

deathBySOFA2 <- aggregate(as.numeric(eICU_Dat$DiedHosp), list(eICU_Dat$SOFAcategory2), mean)
deathBySOFA2[,2] <- round(100*deathBySOFA2[,2], 2)
colnames(deathBySOFA2) <- c("SOFA category", "Mortality")
deathBySOFA2_Asian <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Asian", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "Asian", "SOFAcategory2"]), mean)
deathBySOFA2_Asian[,2] <- round(100*deathBySOFA2_Asian[,2], 2)
colnames(deathBySOFA2_Asian) <- c("SOFA category", "Mortality")
deathBySOFA2_Black <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Black", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "Black", "SOFAcategory2"]), mean)
deathBySOFA2_Black[,2] <- round(100*deathBySOFA2_Black[,2], 2)
colnames(deathBySOFA2_Black) <- c("SOFA category", "Mortality")
deathBySOFA2_White <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "White", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "White", "SOFAcategory2"]), mean)
deathBySOFA2_White[,2] <- round(100*deathBySOFA2_White[,2], 2)
colnames(deathBySOFA2_White) <- c("SOFA category", "Mortality")
deathBySOFA2_Hispanic <- aggregate(as.numeric(eICU_Dat[eICU_Dat$ethnicity == "Hispanic", "DiedHosp"]), list(eICU_Dat[eICU_Dat$ethnicity == "Hispanic", "SOFAcategory2"]), mean)
deathBySOFA2_Hispanic[,2] <- round(100*deathBySOFA2_Hispanic[,2], 2)
colnames(deathBySOFA2_Hispanic) <- c("SOFA category", "Mortality")


theNumbers <- c(nrow(eICU_Dat[eICU_Dat$SOFAcategory2 == 1,]), 
                nrow(eICU_Dat[eICU_Dat$SOFAcategory2 == 2,]), 
                nrow(eICU_Dat[eICU_Dat$SOFAcategory2 == 3,])
              )


deathBySOFA2sum <- cbind(c("0-7","8-11","11+"), theNumbers, deathBySOFA2[,2], deathBySOFA2_Hispanic[,2], deathBySOFA2_Black[,2], deathBySOFA2_White[,2], deathBySOFA2_Asian[,2])
colnames(deathBySOFA2sum) <- c("SOFA category", "Nums", "All", "Hispanic", "Black", "White","Asian" )

SMRbySOFA_sum2 <- data.frame(cbind(deathBySOFA2[,1], theNumbers, matrix( round( as.numeric( deathBySOFA2sum[,3:7]) / as.numeric( deathBySOFA2sum[,3]), 2),3,5)))

colnames(SMRbySOFA_sum2) <- c("SOFA category", "Nums", "All", "Hispanic", "Black", "White", "Asian")

options(OutDec = "\xB7")
knitr::kable(deathBySOFA2sum, format.args = c(width=70), caption = "Mortality rates by ethnic group and SOFA category (eICU).", align='c')

```
 
\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
SMRbySOFA_sum2[,1] <- c("0-7","8-11","11+")
knitr::kable(SMRbySOFA_sum2, format.args = c(width=70), caption = "Mortality in the eICU database by ethnic group and day 1 SOFA category standardised to the total mortality for the SOFA category (eICU).", align='c')

```



```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=figWidth2, fig.height=figHeight2}
# Now the ROC for SOFA scores in eICU
library(ROCR)
thePredictionsAll_SE <- eICU_Dat$SOFA1
thePredictionsAsian_SE <- eICU_Dat[eICU_Dat$ethnicity == "Asian",]$SOFA1
thePredictionsBlack_SE <- eICU_Dat[eICU_Dat$ethnicity == "Black",]$SOFA1
thePredictionsHispanic_SE <- eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$SOFA1
thePredictionsWhite_SE <- eICU_Dat[eICU_Dat$ethnicity == "White",]$SOFA1

theLabelsAll_SE <- eICU_Dat$DiedHosp
theLabelsAsian_SE <- eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp
theLabelsBlack_SE <- eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp
theLabelsHispanic_SE <- eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp
theLabelsWhite_SE <- eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp

predObjAll_SE <- prediction(predictions = thePredictionsAll_SE, labels = theLabelsAll_SE)
aucAll_SE <- round(performance(predObjAll_SE, measure = "auc")@y.values[[1]],3)
perfObjAll_SE <- performance(predObjAll_SE, "tpr","fpr")
predObjAsian_SE <- prediction(predictions = thePredictionsAsian_SE, labels = theLabelsAsian_SE)
aucAsian_SE <- round(performance(predObjAsian_SE, measure = "auc")@y.values[[1]],3)
perfObjAsian_SE <- performance(predObjAsian_SE, "tpr","fpr")
predObjBlack_SE <- prediction(predictions = thePredictionsBlack_SE, labels = theLabelsBlack_SE)
aucBlack_SE <- round(performance(predObjBlack_SE, measure = "auc")@y.values[[1]],3)
perfObjBlack_SE <- performance(predObjBlack_SE, "tpr","fpr")
predObjHispanic_SE <- prediction(predictions = thePredictionsHispanic_SE, labels = theLabelsHispanic_SE)
aucHispanic_SE <- round(performance(predObjHispanic_SE, measure = "auc")@y.values[[1]],3)
perfObjHispanic_SE <- performance(predObjHispanic_SE, "tpr","fpr")
predObjWhite_SE <- prediction(predictions = thePredictionsWhite_SE, labels = theLabelsWhite_SE)
aucWhite_SE <- round(performance(predObjWhite_SE, measure = "auc")@y.values[[1]],3)
perfObjWhite_SE <- performance(predObjWhite_SE, "tpr","fpr")

{
  plot(perfObjAll_SE, main = "ROC curves for the SOFA scores in the eICU \ndatabase by ethnicity.", lwd=3, cex.main=0.9)
  plot(perfObjAsian_SE, add = TRUE, col = 'orange', lwd=3)
  plot(perfObjBlack_SE, add = TRUE, col = 'blue', lwd=3)
  plot(perfObjHispanic_SE, add = TRUE, col = 'red', lwd=3)
  plot(perfObjWhite_SE, add = TRUE, col = 'seagreen4', lwd=3)
  abline(0,1,col='black', lty=2)
  op <- par(cex = 0.8)
  legend('bottomright',legend = c(paste("All", aucAll_SE),paste("Hispanic",aucHispanic_SE),paste("Black", aucBlack_SE),  paste("White",aucWhite_SE),paste("Asian", aucAsian_SE)), col=c("black","red","blue","seagreen4","orange"), lty = c(1,1,1,1,1), lwd=c(3,3,3,3,3))
}

```
 


```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=12, fig.height=8}
# SOFA scores for MIMIC-II
library(pROC)
AUCall_SOFA_M <- round(ci.auc(icustays$DiedHosp, icustays$SOFA_Adm),3)
AUCasian_SOFA_M <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Asian",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Asian",]$SOFA_Adm),3)
AUCblack_SOFA_M <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Black",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Black",]$SOFA_Adm),3)
AUChispanic_SOFA_M <- round(ci.auc(icustays[icustays$ETHNICITY2 == "Hispanic",]$DiedHosp, icustays[icustays$ETHNICITY2 == "Hispanic",]$SOFA_Adm),3)
AUCwhite_SOFA_M <- round(ci.auc(icustays[icustays$ETHNICITY2 == "White",]$DiedHosp, icustays[icustays$ETHNICITY2 == "White",]$SOFA_Adm),3)

plotForest("MIMIC-III",AUChispanic_SOFA_M, AUCblack_SOFA_M, AUCwhite_SOFA_M, AUCasian_SOFA_M, AUCall_SOFA_M, "admission SOFA score", "AUC",doPNG=FALSE, sampleSizes_M, tag="SOFA")

```
 
```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, results='hide', fig.width=12, fig.height=8}
# SOFA score in eICU
AUCall_SOFA_E <- round(ci.auc(eICU_Dat$DiedHosp, eICU_Dat$SOFA1),3)
AUCasian_SOFA_E <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "Asian",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Asian",]$SOFA1),3)
AUCblack_SOFA_E <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "Black",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Black",]$SOFA1),3)
AUChispanic_SOFA_E <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "Hispanic",]$SOFA1),3)
AUCwhite_SOFA_E <- round(ci.auc(eICU_Dat[eICU_Dat$ethnicity == "White",]$DiedHosp, eICU_Dat[eICU_Dat$ethnicity == "White",]$SOFA1),3)

plotForest("eICU", AUChispanic_SOFA_E, AUCblack_SOFA_E, AUCwhite_SOFA_E, AUCasian_SOFA_E, AUCall_SOFA_E, "admission SOFA score", "AUC",doPNG=FALSE, sampleSizes_eICU, tag="SOFA")

```



\pagebreak


## SMRs (lower 95% CO, mean, upper 95% CI)
### SMR, eICU, Apache, all  
`r SMRall_eICU_Ap`  

### SMR, eICU, Apache, Asian  
`r SMRasian_eICU_Ap`  

### SMR, eICU, Apache, Black  
`r SMRblack_eICU_Ap`  

### SMR, eICU, Apache, Hispanic  
`r SMRhispanic_eICU_Ap`  

### SMR, eICU, Apache, White  
`r SMRwhite_eICU_Ap`  

### SMR, MIMIC-III, Apache, all  
`r SMRall_MIMIC_Oa`  

### SMR,  MIMIC-III, Apache, Asian 
`r SMRasian_MIMIC_Oa`  

### SMR,  MIMIC-III, Apache, Black  
`r SMRblack_MIMIC_Oa`  

### SMR,  MIMIC-III, Apache, Hispanic  
`r SMRhispanic_MIMIC_Oa`  

### SMR,  MIMIC-III, Apache, White  
`r SMRwhite_MIMIC_Oa`

  

\pagebreak


## Counts

The total number of episodes in the MIMIC-II data is `r MIMICnos$Total`.

Once all occurrences of reported ages below `r minAge` and above `r maxAge` are removed is `r MIMICnos$exAges`.

Of these, `r MIMICnos$incEthnicity1` are in the list of included ethnic groups (`r ethnicitiesList`).


```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
SMRcalc <- data.frame(Ethnicity = character(), 
                 Source = character(), 
                 Count = integer(), 
                 "Predicted mortality" = numeric(), 
                 "Actual mortality" = numeric(),
                 "Predicted deaths" = integer(),
                 "Actual death" = integer()
                 )
for(eth in ethnicitiesList){
  
  if(eth == "Hispanic"){
    eth2 <- "Hispanic"
  }else if(eth == "Asian"){
    eth2 <- "Asian"
  }else if(eth == "White"){
    eth2 <- "White"
  }else{
    eth2 <- "Black"
  }
  
  
  theRow <- c(eth, 
              "eICU", 
              nrow(eICU_Dat[eICU_Dat$ethnicity == eth, ]),
              round(mean(eICU_Dat[eICU_Dat$ethnicity == eth, ]$apacheMort),2),
              round(mean(eICU_Dat[eICU_Dat$ethnicity == eth, ]$apacheMort) * nrow(eICU_Dat[eICU_Dat$ethnicity == eth, ]),0),
              nrow(eICU_Dat[eICU_Dat$ethnicity == eth & eICU_Dat$DiedHosp == TRUE, ]),
              round(mean(eICU_Dat[eICU_Dat$ethnicity == eth, ]$DiedHosp) / mean(eICU_Dat[eICU_Dat$ethnicity == eth, ]$apacheMort),2)
              )
  theRow2 <- c(eth, 
              "MIMIC-III", 
              nrow(icustays[icustays$ETHNICITY2 == eth2, ]),
              round(mean(icustays[icustays$ETHNICITY2 == eth2, ]$OASISprob),2),
              round(mean(icustays[icustays$ETHNICITY2 == eth2, ]$OASISprob) * nrow(icustays[icustays$ETHNICITY2 == eth2, ]),0),
              nrow(icustays[icustays$ETHNICITY2 == eth2 & icustays$DiedHosp == TRUE, ]),
              round(mean(icustays[icustays$ETHNICITY2 == eth2, ]$DiedHosp) / mean(icustays[icustays$ETHNICITY2 == eth2, ]$OASISprob),2)
              )
  
  SMRcalc <- rbind(SMRcalc, theRow, theRow2)
}

colnames(SMRcalc) <- c("Ethnicity", "Dataset", "Total number", "Mean predicted mortality","Predicted deaths", "Actual deaths", "SMR")

knitr::kable(SMRcalc, format.args = c(width=70), caption = "Predicted and observed mortality in different ethnic groups along (SMR = actual/predicted mortality ratio).", align='c')

```



```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
SOFO_0_M <- nrow(icustays[icustays$SOFA_Adm == 0,])
SOFO_0_E <- nrow(eICU_Dat[eICU_Dat$SOFA1 == 0,])

SOFO_0_M_D <- nrow(icustays[icustays$SOFA_Adm == 0 & icustays$DiedHosp == TRUE,])
SOFO_0_E_D <- nrow(eICU_Dat[eICU_Dat$SOFA1 == 0 & eICU_Dat$DiedHosp == 1,])

```

There were `r SOFO_0_E` admissions with an an initial SOFA score of 0 in the eICU database, of which, `r SOFO_0_E_D` died.
There were `r SOFO_0_M` admissions with an an initial SOFA score of 0 in the MIMIC-III database, of which, `r SOFO_0_M_D` died.

\pagebreak

# Explanatory power of ethnicity and predictive scoring

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
LR_Apache <- glm(DiedHosp ~ apachescore, data = eICU_Dat,family=binomial(link=logit))
R2_Apache <- rsq(LR_Apache)
knitr::kable(round(summary(LR_Apache)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of Apache score and in-hospital death (eICU)", align = 'c')
```
R^2^ = `r R2_Apache`.

\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
LR_OASIS <- glm(DiedHosp ~ OASIS, data = icustays,family=binomial(link=logit))
R2_Oasis <- rsq(LR_OASIS)
knitr::kable(round(summary(LR_OASIS)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of Oasis score and in-hospital death (MIMIC-III)", align = 'c')
```
R^2^ = `r R2_Oasis`.

\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
LR_SOFA_MIMIC <- glm(DiedHosp ~ SOFA_Adm, data = icustays,family=binomial(link=logit))
R2_SOFA_MIMIC <- rsq(LR_SOFA_MIMIC)
knitr::kable(round(summary(LR_SOFA_MIMIC)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of SOFA score and in-hospital death (MIMIC-III)", align = 'c')
```
R^2^ = `r R2_SOFA_MIMIC`.

\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
LR_SOFA_eICU <- glm(DiedHosp ~ SOFA1, data = eICU_Dat,family=binomial(link=logit))
R2_SOFA_eICU <- rsq(LR_SOFA_eICU)
knitr::kable(round(summary(LR_SOFA_eICU)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of SOFA score and in-hospital death (eICU)", align = 'c')
```
R^2^ = `r R2_SOFA_eICU`.

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

LR_MIMIC <- glm(DiedHosp ~ ETHNICITY2, data = icustays,family=binomial(link=logit))
R2_MIMIC <- rsq(LR_MIMIC)
LR_eICU <- glm(DiedHosp ~ ethnicity, data = eICU_Dat,family=binomial(link=logit))
R2_eICU <- rsq(LR_eICU)

knitr::kable(round(summary(LR_eICU)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of ethnicity and in-hospital death (eICU)", align='c')

```
R^2^ = `r R2_eICU`.

\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
knitr::kable(round(summary(LR_MIMIC)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of ethnicity and in-hospital death (MIMIC-III)", align='c')

```
R^2^ = `r R2_MIMIC`.

\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
LR_MIMIC_Oasis <- glm(DiedHosp ~ ETHNICITY2 + OASIS, data = icustays,family=binomial(link=logit))
R2_MIMIC_Oasis <- rsq(LR_MIMIC_Oasis)
LR_eICU_Apache <- glm(DiedHosp ~ ethnicity + apachescore, data = eICU_Dat,family=binomial(link=logit))
R2_eICU_Apache <- rsq(LR_eICU_Apache)

knitr::kable(round(summary(LR_eICU_Apache)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of ethnicity and in-hospital death controlled for Apache score (eICU).", align='c')

```
R^2^ = `r R2_eICU_Apache`.

\pagebreak

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

knitr::kable(round(summary(LR_MIMIC_Oasis)$coefficients,3), format.args = c(width=70), caption = "Logistic regression of ethnicity and in-hospital death controlled for Oasis score (MIMIC-II).", align='c')

```
R^2^ = `r R2_MIMIC_Oasis`.

\

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}

LR_SOFA_eICU_Eth <- glm(DiedHosp ~ ethnicity + SOFA1, data = eICU_Dat,family=binomial(link=logit))
R2_SOFA_eICU_Eth <- rsq(LR_SOFA_eICU_Eth)

knitr::kable(round(summary(LR_SOFA_eICU_Eth)$coefficients, 3), format.args = c(width=70), caption = "Logistic regression of ethnicity and in-hospital death controlled for SOFA score (eICU).", align='c')

```
R^2^ = `r R2_SOFA_eICU`.

\


R^2^ = `r R2_SOFA_MIMIC_Eth`.

```{r eval = TRUE, echo = FALSE, warning=FALSE, message = FALSE}
R2_Partial_SOFA_eICU <- rsq.partial(LR_SOFA_eICU_Eth)
R2_Partial_Apache_Eth <- rsq.partial(LR_eICU_Apache)

 R2_Partial_SOFA_MIMIC <- rsq.partial(LR_SOFA_MIMIC_Eth)
R2_Partial_Oasis_Eth <- rsq.partial(LR_MIMIC_Oasis)

```

\pagebreak

# The contibution of the risk scores and ethnicity to variation in in-hospital mortality in the eICU and MIMIC-II databases.

The Oasis score explains `r round(R2_Partial_Oasis_Eth$partial.rsq[2]*100, 3)`% of the variation in mortality, and the ethnicity explains `r round(R2_Partial_Oasis_Eth$partial.rsq[1]*100, 3)`%.  

The Apache score explains `r round(R2_Partial_Apache_Eth$partial.rsq[2]*100, 3)`% of the variation in mortality, and the ethnicity explains `r round(R2_Partial_Apache_Eth$partial.rsq[1]*100, 3)`%.  

The SOFA score explains `r round(R2_Partial_SOFA_MIMIC$partial.rsq[2]*100, 3)`% of the variation in mortality, and the ethnicity explains `r round(R2_Partial_SOFA_MIMIC$partial.rsq[1]*100, 3)`% in the MIMIC-II database.  

The SOFA score explains `r round(R2_Partial_SOFA_eICU$partial.rsq[2]*100, 3)`% of the variation in mortality, and the ethnicity explains `r round(R2_Partial_SOFA_eICU$partial.rsq[1]*100, 3)`% in the eICU database.  
